---
title: "Living Review: Long COVID & ME/CFS"
subtitle: "A living systematic review and meta-analysis"
toc: false
number-sections: false
---

```{r}
#| label: setup
#| include: false
library(targets)
tar_config_set(store = here::here("_targets"))
source("R/utils.R")
studies <- targets::tar_read(studies)
```

::: {.lead}
This is a **living systematic review** of research on Long COVID (post-COVID condition) and Myalgic Encephalomyelitis/Chronic Fatigue Syndrome (ME/CFS). Updated automatically as new research is published, with periodic manual verification and data extraction.
:::

## Current Status

```{r}
#| label: status-metrics
#| results: asis
#| echo: false
if (nrow(studies) > 0) {
  n_total <- nrow(studies)
  n_included <- sum(studies$included, na.rm = TRUE)
  n_pending <- sum(!studies$human_verified, na.rm = TRUE)
  n_extracted <- sum(studies$data_extracted, na.rm = TRUE)

  cat('<div class="status-grid">\n')
  cat(sprintf('<div class="status-metric"><span class="metric-value">%d</span><span class="metric-label">Studies identified</span></div>\n', n_total))
  cat(sprintf('<div class="status-metric"><span class="metric-value">%d</span><span class="metric-label">Included</span></div>\n', n_included))
  cat(sprintf('<div class="status-metric"><span class="metric-value">%d</span><span class="metric-label">Pending review</span></div>\n', n_pending))
  cat(sprintf('<div class="status-metric"><span class="metric-value">%d</span><span class="metric-label">Data extracted</span></div>\n', n_extracted))
  cat('</div>\n')
} else {
  cat("*No studies in database yet. Run the search pipeline to populate.*\n")
}
```

## Review Domains

::: {.grid}

::: {.g-col-6}
::: {.domain-card .domain-card--treatment}
### Treatment Effectiveness

Comparative effectiveness of interventions including pacing, graded exercise therapy, CBT, pharmacological treatments, and supplements.

[Explore treatment evidence](reviews/treatment-effectiveness.qmd)
:::
:::

::: {.g-col-6}
::: {.domain-card .domain-card--prevalence}
### Prevalence

Prevalence estimates for Long COVID and ME/CFS symptoms across populations and diagnostic criteria.

[View prevalence data](reviews/prevalence.qmd)
:::
:::

::: {.g-col-6}
::: {.domain-card .domain-card--pem}
### PEM Prevalence

Prevalence and characterisation of post-exertional malaise — the hallmark symptom of ME/CFS.

[Explore PEM evidence](reviews/pem-prevalence.qmd)
:::
:::

::: {.g-col-6}
::: {.domain-card .domain-card--recovery}
### Recovery Trajectories

Natural history and recovery patterns, stratified by PEM status and condition severity.

[View recovery data](reviews/recovery-trajectories.qmd)
:::
:::

:::

Clinical studies and patient-reported outcome studies are included but analysed separately within each domain.

## How It Works {.unnumbered}

::: {.pipeline-flow}
::: {.pipeline-step}
[🔍]{.step-icon}
[Automated Search]{.step-label}
[Weekly Europe PMC queries]{.step-detail}
:::
::: {.pipeline-step}
[🤖]{.step-icon}
[LLM Screening]{.step-label}
[Relevance classification]{.step-detail}
:::
::: {.pipeline-step}
[👁]{.step-icon}
[Human Verification]{.step-label}
[Manual review of flagged studies]{.step-detail}
:::
::: {.pipeline-step}
[📊]{.step-icon}
[Data Extraction]{.step-label}
[Structured outcome data]{.step-detail}
:::
::: {.pipeline-step}
[📈]{.step-icon}
[Meta-Analysis]{.step-label}
[Pooled estimates & plots]{.step-detail}
:::
::: {.pipeline-step}
[🌐]{.step-icon}
[Publication]{.step-label}
[Auto-rendered Quarto site]{.step-detail}
:::
:::

See the [Methods](methods.qmd) page for full details on search strategies, inclusion criteria, data extraction, and quality assessment.
