---
title: "Living Review: Long COVID & ME/CFS"
subtitle: "A living systematic review and meta-analysis"
---

```{r}
#| label: setup
#| include: false
source("R/utils.R")
studies <- read_study_database()
```

## About This Review

This is a **living systematic review** of research on Long COVID (post-COVID condition) and Myalgic Encephalomyelitis/Chronic Fatigue Syndrome (ME/CFS). It is updated automatically as new research is published, with periodic manual verification and data extraction.

The review covers four domains:

1. **[Treatment Effectiveness](reviews/treatment-effectiveness.qmd)** — Comparative effectiveness of interventions including pacing, graded exercise therapy, CBT, pharmacological treatments, and supplements
2. **[Prevalence](reviews/prevalence.qmd)** — Prevalence estimates for Long COVID and ME/CFS symptoms
3. **[PEM Prevalence](reviews/pem-prevalence.qmd)** — Prevalence and characterisation of post-exertional malaise
4. **[Recovery Trajectories](reviews/recovery-trajectories.qmd)** — Natural history and recovery patterns, stratified by PEM status

Clinical studies and patient-reported outcome studies are included but analysed separately within each domain.

## Current Status

```{r}
#| label: status-summary
if (nrow(studies) > 0) {
  n_total <- nrow(studies)
  n_included <- sum(studies$included, na.rm = TRUE)
  n_pending <- sum(!studies$human_verified, na.rm = TRUE)
  n_extracted <- sum(studies$data_extracted, na.rm = TRUE)
  date_range <- range(studies$date_added, na.rm = TRUE)

  cat(sprintf(
    "- **Studies identified**: %d\n- **Studies included**: %d\n- **Pending review**: %d\n- **Data extracted**: %d\n- **Database period**: %s to %s\n",
    n_total, n_included, n_pending, n_extracted,
    date_range[1], date_range[2]
  ))
} else {
  cat("*No studies in database yet. Run the search pipeline to populate.*\n")
}
```

## Methods

See the [Methods](methods.qmd) page for search strategies, inclusion/exclusion criteria, data extraction procedures, and quality assessment framework.

## How It Works

```{mermaid}
flowchart LR
    A[Automated\nPubMed Search] --> B[LLM\nScreening]
    B --> C[Human\nVerification]
    C --> D[Data\nExtraction]
    D --> E[Meta-Analysis\n& Plots]
    E --> F[Quarto\nWebsite]
```

1. **Automated search**: GitHub Actions runs weekly PubMed/Europe PMC searches using predefined queries
2. **LLM screening**: New results are screened by an LLM for relevance against inclusion criteria
3. **Human verification**: Flagged studies are reviewed manually
4. **Data extraction**: Included studies have outcomes extracted to structured CSVs
5. **Analysis & publication**: Meta-analyses re-run and the website re-renders automatically
