---
title: "Study Review Tool"
---

```{r}
#| label: setup
#| include: false
library(targets)
tar_config_set(store = here::here("_targets"))
source(here::here("R/utils.R"))
studies <- targets::tar_read(studies)
```

The review tool is a Shiny application for screening studies and extracting data. It reads and writes to the same `studies.parquet` database that powers this website.

```{r}
#| label: review-status
#| results: asis
if (nrow(studies) > 0) {
  n_llm_relevant <- sum(studies$llm_relevant %in% TRUE & !studies$human_verified %in% TRUE)
  n_unscreened <- sum(is.na(studies$llm_relevant) & !studies$human_verified %in% TRUE)
  n_included_no_extract <- sum(studies$included %in% TRUE & !studies$data_extracted %in% TRUE)

  cli::cat_line("## Pending Work\n")
  cli::cat_line(glue::glue("- **LLM-flagged, needs human review**: {n_llm_relevant}"))
  cli::cat_line(glue::glue("- **Unscreened**: {n_unscreened}"))
  cli::cat_line(glue::glue("- **Included, needs data extraction**: {n_included_no_extract}"))
}
```

## Running Locally

```bash
Rscript -e 'shiny::runApp("app-review.R")'
```

## Deployment

The app is deployed separately from this static site. After reviewing or extracting data, commit the updated parquet files and the website re-renders automatically.

::: {.callout-tip}
## Deploy to shinyapps.io

```r
rsconnect::deployApp(
  appDir = ".",
  appFiles = c("app-review.R", "R/utils.R", "R/extraction.R", "data/studies.parquet"),
  appName = "lc-review"
)
```

Once deployed, update the link below.
:::

::: {.callout-note}
## App link
The review tool is not yet deployed. Run it locally with the command above, or deploy to [shinyapps.io](https://www.shinyapps.io/) / Posit Connect and update this page with the URL.
:::
