---
title: "Study Database"
---

```{r}
#| label: setup
#| include: false
source("R/utils.R")
studies <- read_study_database()
has_data <- nrow(studies) > 0
```

## All Studies

Interactive table of all studies in the database, with filtering and search.

```{r}
#| label: study-table
#| eval: !expr has_data
if (has_data) {
  display <- studies[, c(
    "study_id", "year", "title", "journal",
    "search_topic", "included", "human_verified",
    "study_type", "reporting_type", "data_extracted", "date_added"
  )]

  DT::datatable(
    display,
    filter = "top",
    options = list(
      pageLength = 25,
      scrollX = TRUE,
      order = list(list(10, "desc"))
    ),
    rownames = FALSE
  )
} else {
  cat("*No studies in database yet.*\n")
}
```

## PRISMA Flow

```{r}
#| label: prisma-flow
#| eval: !expr has_data
if (has_data) {
  n_identified <- nrow(studies)
  n_screened <- sum(studies$llm_relevant | studies$human_verified, na.rm = TRUE)
  n_excluded_screen <- n_identified - n_screened
  n_assessed <- sum(studies$human_verified, na.rm = TRUE)
  n_included <- sum(studies$included, na.rm = TRUE)
  n_excluded_full <- n_assessed - n_included
  n_extracted <- sum(studies$data_extracted, na.rm = TRUE)
}
```

```{mermaid}
%%| eval: !expr has_data
flowchart TD
    A[Studies identified\nthrough search] --> B[LLM screening]
    B --> C[Potentially relevant]
    B --> D[Excluded at screening]
    C --> E[Human full-text review]
    E --> F[Included in review]
    E --> G[Excluded at full-text]
    F --> H[Data extracted]
```

## Search Log

```{r}
#| label: search-log
log_path <- "data/search-log.csv"
if (file.exists(log_path)) {
  log <- read.csv(log_path, stringsAsFactors = FALSE)
  if (nrow(log) > 0) {
    DT::datatable(
      log[, c("date", "topic", "n_results")],
      options = list(pageLength = 20, order = list(list(0, "desc"))),
      rownames = FALSE
    )
  } else {
    cat("*No searches logged yet.*\n")
  }
} else {
  cat("*No search log found.*\n")
}
```

## Database Schema

The study database tracks each study through the pipeline:

| Field | Description |
|-------|-------------|
| `study_id` | Unique identifier (first_author + year) |
| `pmid`, `doi` | PubMed ID and DOI |
| `date_added` | Date study entered the database |
| `llm_relevant` | Initial LLM relevance screening result |
| `llm_category` | LLM-assigned review domain |
| `human_verified` | Whether a human has reviewed the study |
| `included` | Final inclusion decision |
| `reporting_type` | Clinical or patient-reported |
| `pem_assessed` | Whether PEM was assessed in the study |
| `data_extracted` | Whether outcome data has been extracted |
