---
title: "Biomarkers"
subtitle: "Biological markers in Long COVID and ME/CFS research"
aliases:
  - /reviews/biomarkers/
---

```{r}
#| label: setup
#| include: false
library(targets)
tar_config_set(store = here::here("_targets"))
source(here::here("R/utils.R"))
source(here::here("R/meta-analysis.R"))
source(here::here("R/plots.R"))

studies <- targets::tar_read(studies)
bm <- targets::tar_read(biomarkers)
has_data <- nrow(bm) > 0
```

::: {.lead}
Biomarkers are measurable biological signals -- proteins, metabolites, immune markers, imaging findings -- that can reveal what is happening in the body at a level symptoms alone cannot reach.
For Long COVID and ME/CFS, biomarker research serves two purposes: it can validate patients' experiences with objective evidence, and it can point toward treatment targets.
This review synthesises biomarker findings across immune dysfunction, autoantibodies, neuroinflammation, microclots, endothelial dysfunction, metabolomics, and proteomics.
:::

```{r}
#| label: included-studies
#| results: asis
#| echo: false
topic_studies <- studies |>
  dplyr::filter(
    !is.na(search_topic), search_topic == "biomarkers",
    !is.na(llm_relevant), llm_relevant
  )
if (nrow(topic_studies) > 0) {
  cat(sprintf('<div class="study-count"><span class="count-number">%d</span><span class="count-label">relevant<br>studies</span></div>\n', nrow(topic_studies)))
}
```

::: {.at-a-glance}

## At a Glance {#sec-lay}

Researchers are searching for biological markers — measurable signals in blood, brain scans, and other tests — that could explain what is going wrong in ME/CFS and Long COVID.
Most findings so far point to immune system problems, but the evidence is still early.

```{r}
#| label: lay-biomarker-types
#| eval: !expr has_data
#| fig-height: 5
#| fig-alt: "Horizontal bar chart showing the number of biomarker findings by biological category."
#| fig-cap: "Where are researchers looking? Each bar shows how many findings have been reported for that type of biological marker. Immune markers and cytokines dominate, but this reflects research focus, not necessarily where the most important signals are."
if (has_data && "biomarker_type" %in% names(bm)) {
  type_counts <- bm |>
    dplyr::filter(!is.na(biomarker_type)) |>
    dplyr::count(biomarker_type, sort = TRUE) |>
    dplyr::mutate(biomarker_type = factor(biomarker_type, levels = rev(biomarker_type)))
  if (nrow(type_counts) > 0) {
    ggplot(type_counts, aes(x = n, y = biomarker_type)) +
      geom_col(fill = cols_review["clinical"], width = 0.7) +
      geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
      scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
      labs(x = "Number of findings", y = NULL,
           title = "What biological markers are researchers studying?",
           subtitle = "Number of findings by marker type") +
      theme_lay
  }
}
```

```{r}
#| label: lay-biomarker-direction
#| eval: !expr has_data
#| fig-height: 4
#| fig-alt: "Stacked bar chart showing what proportion of biomarker findings show elevated, reduced, mixed, or no difference results."
#| fig-cap: "What do the biomarker studies find? Most findings show markers are either elevated (higher than normal) or give mixed results across studies. 'Mixed' means different studies disagree — a sign that more research is needed."
if (has_data && "finding_direction" %in% names(bm)) {
  dir_counts <- bm |>
    dplyr::filter(!is.na(finding_direction)) |>
    dplyr::count(finding_direction)
  if (nrow(dir_counts) > 0) {
    dir_counts$pct <- dir_counts$n / sum(dir_counts$n)
    dir_colours <- c("elevated" = "#D55E00", "reduced" = "#0072B2",
                     "no difference" = "#999999", "mixed" = "#E69F00")
    ggplot(dir_counts, aes(x = n, y = "Findings", fill = finding_direction)) +
      geom_col(position = "stack", width = 0.6) +
      geom_text(aes(label = paste0(n, " (", round(pct * 100), "%)")),
                position = position_stack(vjust = 0.5), size = 3.5,
                colour = "white", fontface = "bold") +
      scale_fill_manual(values = dir_colours, name = NULL) +
      labs(x = "Number of findings", y = NULL,
           title = "What do researchers find?",
           subtitle = "Direction of biomarker changes in patients vs healthy controls") +
      theme_lay +
      theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
            legend.position = "bottom")
  }
}
```

:::

<hr class="section-divider">

## Biomarker Types & Directions {.tabset}

::: {.panel-tabset}

### Biomarker Types

The table below groups every extracted biomarker finding by its biological category.
Some categories (immune markers, cytokines) appear far more often than others, reflecting where researchers have focused -- not necessarily where the strongest signals live.
Gaps in underrepresented categories may be as informative as the clusters of findings in well-studied ones.

```{r}
#| label: type-summary
#| eval: !expr has_data
if (has_data && "biomarker_type" %in% names(bm)) {
  type_counts <- bm |>
    dplyr::filter(!is.na(biomarker_type)) |>
    dplyr::count(biomarker_type, sort = TRUE)
  if (nrow(type_counts) > 0) {
    names(type_counts) <- c("Biomarker Type", "Findings")
    display_table(type_counts)
  }
}
```

### Direction of Findings

Whether a biomarker is elevated or reduced in patients relative to controls matters enormously for interpretation.
A consistently elevated inflammatory marker across many studies strengthens the case for that pathway's involvement.
Mixed results -- some studies finding elevation, others finding no difference -- may reflect genuine biological heterogeneity across patient subgroups, or they may signal measurement inconsistencies.
The chart below stacks finding directions for each biomarker type so you can see at a glance where the evidence converges and where it diverges.

```{r}
#| label: direction-data
#| include: false
dir_counts <- NULL
dir_h <- 6
if (has_data && all(c("biomarker_type", "finding_direction") %in% names(bm))) {
  dir_df <- bm |>
    dplyr::filter(!is.na(biomarker_type), !is.na(finding_direction))
  if (nrow(dir_df) > 0) {
    dir_counts <- dir_df |>
      dplyr::count(Type = biomarker_type, Direction = finding_direction) |>
      dplyr::filter(n > 0)
    dir_h <- bar_height(dplyr::n_distinct(dir_counts$Type))
  }
}
```

```{r}
#| label: direction-plot
#| eval: !expr '!is.null(dir_counts)'
#| fig-height: !expr dir_h
#| fig-cap: "Stacked bar chart showing how many findings report each biomarker type as elevated, reduced, mixed, or showing no difference -- revealing where evidence converges."
#| fig-alt: "Horizontal stacked bar chart with biomarker types on the y-axis and number of findings on the x-axis. Each bar is segmented by finding direction: elevated (orange-red), reduced (blue), no difference (grey), and mixed (yellow)."
library(ggplot2)
ggplot(dir_counts, aes(x = reorder(Type, n), y = n, fill = Direction)) +
  geom_col(position = "stack") +
  coord_flip() +
  scale_fill_manual(
    values = c(elevated = "#D55E00", reduced = "#0072B2",
               "no difference" = "#999999", mixed = "#E69F00"),
    name = "Direction"
  ) +
  labs(x = NULL, y = "Number of findings") +
  theme_review
```

:::

## Condition Comparison {#sec-condition}

Long COVID and ME/CFS share substantial symptom overlap, and a key question is whether they also share biological mechanisms.

::: {.key-finding}
When a biomarker finding appears in studies of both conditions, it strengthens the case for shared pathophysiology.
Findings unique to one condition may reflect genuine biological differences or simply gaps in the research -- ME/CFS has decades of biomarker literature, while Long COVID research is only a few years old.
:::

The table below shows how findings distribute across conditions.

```{r}
#| label: condition-comparison
#| eval: !expr has_data
if (has_data && all(c("condition", "biomarker_name", "finding_direction") %in% names(bm))) {
  both_conditions <- bm |> dplyr::filter(condition == "both")
  if (nrow(both_conditions) > 0) {
    cli::cat_line(glue::glue("**{nrow(both_conditions)}** findings reported for both Long COVID and ME/CFS simultaneously."))
  }
  cond_counts <- bm |>
    dplyr::filter(!is.na(condition)) |>
    dplyr::count(condition)
  names(cond_counts) <- c("Condition", "Findings")
  display_table(cond_counts)
}
```

## Top Biomarkers {#sec-top}

The most-studied biomarkers are listed here ranked by how many separate study findings mention them.
High frequency does not equal high importance -- it often reflects ease of measurement or existing research momentum.
Still, biomarkers that appear repeatedly across independent groups are worth watching closely.

```{r}
#| label: top-biomarkers
#| eval: !expr has_data
if (has_data && "biomarker_name" %in% names(bm)) {
  top20 <- bm |>
    dplyr::filter(!is.na(biomarker_name)) |>
    dplyr::count(biomarker_name, sort = TRUE) |>
    dplyr::slice_head(n = 20)
  names(top20) <- c("Biomarker", "Studies")
  display_table(top20)
}
```

## Diagnostic Performance {#sec-diagnostic}

A biomarker is only clinically useful if it can reliably distinguish patients from healthy controls.
Sensitivity tells you how often the test correctly identifies someone who has the condition; specificity tells you how often it correctly rules out someone who does not.
Few biomarkers for Long COVID or ME/CFS have reached the threshold for clinical diagnostic use, but the table below shows what has been reported so far.

```{r}
#| label: diagnostic-table
#| eval: !expr has_data
if (has_data && all(c("sensitivity", "specificity") %in% names(bm))) {
  diag_df <- bm |>
    dplyr::filter(!is.na(sensitivity), !is.na(specificity))
  if (nrow(diag_df) > 0) {
    display <- diag_df |>
      dplyr::select(dplyr::any_of(c("study_id", "biomarker_name", "condition",
                                    "sensitivity", "specificity"))) |>
      dplyr::mutate(dplyr::across(c(sensitivity, specificity), \(x) round(x, 3)))
    names(display) <- gsub("_", " ", names(display)) |> tools::toTitleCase()
    display_table(display)
  } else {
    cli::cat_line("No sensitivity/specificity data extracted yet.")
  }
}
```

## PEM & Biomarkers {#sec-pem}

Post-exertional malaise (PEM) -- the hallmark worsening of symptoms after physical or cognitive effort -- is the defining feature of ME/CFS and appears in a large subset of Long COVID patients.
Whether a biomarker study assessed PEM in its participants matters: a sample that includes patients with PEM may show different biological profiles than one that does not.
Tracking how often PEM is assessed alongside biomarker measurement helps us judge the quality and applicability of findings.

```{r}
#| label: pem-biomarkers
#| results: asis
#| eval: !expr has_data
if (has_data && "pem_assessed" %in% names(bm)) {
  n_pem <- sum(bm$pem_assessed, na.rm = TRUE)
  pct <- round(n_pem / nrow(bm) * 100)
  cli::cat_line(glue::glue("PEM was assessed alongside biomarkers in **{n_pem}/{nrow(bm)}** ({pct}%) findings."))
}
```

::: {.callout-note collapse="true"}
## Living Review

This page is part of a living systematic review and is updated as new studies are identified, screened, and extracted.
The data, figures, and pooled estimates will change over time.

```{r}
#| label: data-status
#| results: asis
if (!has_data) {
  cli::cat_line("No extracted data yet. Run the extraction pipeline to populate this review.")
} else {
  n_studies <- dplyr::n_distinct(bm$study_id)
  cli::cat_line(glue::glue("{nrow(bm)} biomarker findings from {n_studies} studies."))
}
```
:::
