---
title: "Biomarkers"
aliases:
  - /reviews/biomarkers/
---

```{r}
#| label: setup
#| include: false
source(here::here("R/utils.R"))
source(here::here("R/meta-analysis.R"))
source(here::here("R/plots.R"))

studies <- read_study_database()
bm <- read_extracted_data("biomarkers")
has_data <- nrow(bm) > 0
```

## Overview

This review synthesises biomarker findings in Long COVID and ME/CFS, covering immune dysfunction, autoantibodies, neuroinflammation, microclots, endothelial dysfunction, metabolomics, and proteomics.

```{r}
#| label: included-studies
#| results: asis
topic_studies <- studies |>
  dplyr::filter(
    !is.na(search_topic), search_topic == "biomarkers",
    !is.na(llm_relevant), llm_relevant
  )
if (nrow(topic_studies) > 0) {
  cli::cat_line(glue::glue("**{nrow(topic_studies)} relevant studies** identified."))
}
```

## Biomarker Types {#sec-types}

```{r}
#| label: type-summary
#| eval: !expr has_data
if (has_data && "biomarker_type" %in% names(bm)) {
  type_counts <- bm |>
    dplyr::filter(!is.na(biomarker_type)) |>
    dplyr::count(biomarker_type, sort = TRUE)
  if (nrow(type_counts) > 0) {
    names(type_counts) <- c("Biomarker Type", "Findings")
    display_table(type_counts)
  }
}
```

## Direction of Findings {#sec-direction}

```{r}
#| label: direction-data
#| include: false
dir_counts <- NULL
dir_h <- 6
if (has_data && all(c("biomarker_type", "finding_direction") %in% names(bm))) {
  dir_df <- bm |>
    dplyr::filter(!is.na(biomarker_type), !is.na(finding_direction))
  if (nrow(dir_df) > 0) {
    dir_counts <- dir_df |>
      dplyr::count(Type = biomarker_type, Direction = finding_direction) |>
      dplyr::filter(n > 0)
    dir_h <- bar_height(dplyr::n_distinct(dir_counts$Type))
  }
}
```

```{r}
#| label: direction-plot
#| eval: !expr '!is.null(dir_counts)'
#| fig-height: !expr dir_h
#| fig-cap: "Direction of biomarker findings by type"
library(ggplot2)
ggplot(dir_counts, aes(x = reorder(Type, n), y = n, fill = Direction)) +
  geom_col(position = "stack") +
  coord_flip() +
  scale_fill_manual(
    values = c(elevated = "#D55E00", reduced = "#0072B2",
               "no difference" = "#999999", mixed = "#E69F00"),
    name = "Direction"
  ) +
  labs(x = NULL, y = "Number of findings") +
  theme_review
```

## Condition Comparison {#sec-condition}

```{r}
#| label: condition-comparison
#| eval: !expr has_data
if (has_data && all(c("condition", "biomarker_name", "finding_direction") %in% names(bm))) {
  both_conditions <- bm |> dplyr::filter(condition == "both")
  if (nrow(both_conditions) > 0) {
    cli::cat_line(glue::glue("**{nrow(both_conditions)}** findings reported for both Long COVID and ME/CFS simultaneously."))
  }
  cond_counts <- bm |>
    dplyr::filter(!is.na(condition)) |>
    dplyr::count(condition)
  names(cond_counts) <- c("Condition", "Findings")
  display_table(cond_counts)
}
```

## Top Biomarkers {#sec-top}

```{r}
#| label: top-biomarkers
#| eval: !expr has_data
if (has_data && "biomarker_name" %in% names(bm)) {
  top20 <- bm |>
    dplyr::filter(!is.na(biomarker_name)) |>
    dplyr::count(biomarker_name, sort = TRUE) |>
    dplyr::slice_head(n = 20)
  names(top20) <- c("Biomarker", "Studies")
  display_table(top20)
}
```

## Diagnostic Performance {#sec-diagnostic}

```{r}
#| label: diagnostic-table
#| eval: !expr has_data
if (has_data && all(c("sensitivity", "specificity") %in% names(bm))) {
  diag_df <- bm |>
    dplyr::filter(!is.na(sensitivity), !is.na(specificity))
  if (nrow(diag_df) > 0) {
    display <- diag_df |>
      dplyr::select(dplyr::any_of(c("study_id", "biomarker_name", "condition",
                                    "sensitivity", "specificity"))) |>
      dplyr::mutate(dplyr::across(c(sensitivity, specificity), \(x) round(x, 3)))
    names(display) <- gsub("_", " ", names(display)) |> tools::toTitleCase()
    display_table(display)
  } else {
    cli::cat_line("No sensitivity/specificity data extracted yet.")
  }
}
```

## PEM & Biomarkers {#sec-pem}

```{r}
#| label: pem-biomarkers
#| results: asis
#| eval: !expr has_data
if (has_data && "pem_assessed" %in% names(bm)) {
  n_pem <- sum(bm$pem_assessed, na.rm = TRUE)
  pct <- round(n_pem / nrow(bm) * 100)
  cli::cat_line(glue::glue("PEM was assessed alongside biomarkers in **{n_pem}/{nrow(bm)}** ({pct}%) findings."))
}
```

::: {.callout-note}
## Data Status

```{r}
#| label: data-status
#| results: asis
if (!has_data) {
  cli::cat_line("No extracted data yet. Run the extraction pipeline to populate this review.")
} else {
  n_studies <- dplyr::n_distinct(bm$study_id)
  cli::cat_line(glue::glue("{nrow(bm)} biomarker findings from {n_studies} studies."))
}
```
:::
