---
title: "GRADE Evidence Profiles"
aliases:
  - /reviews/grade/
---

## What is GRADE?

GRADE (Grading of Recommendations, Assessment, Development and Evaluation) is a systematic framework for rating the certainty of evidence.
It answers a specific question: how confident should we be that the estimated effect is close to the true effect?
Certainty is not the same as importance -- a finding can be critically important even if the evidence is rated "very low."

Each outcome starts at "high" certainty and can be downgraded for five reasons: risk of bias in the studies, inconsistency across results, indirectness (the studies do not quite match the question), imprecision (wide confidence intervals), and publication bias.
It can also be upgraded if there is a large effect, a dose-response gradient, or if plausible confounders would have reduced the observed effect.

## How to Read the Table

The table below groups outcomes by research domain.
The **Certainty** column is colour-coded:

- **High** (green): further research is very unlikely to change the estimate.
- **Moderate** (amber): further research is likely to have an important impact and may change the estimate.
- **Low** (orange): further research is very likely to change the estimate.
- **Very low** (red): the estimate is very uncertain.

For clinical decision-making, "very low" certainty does not mean the finding is wrong -- it means we need more and better studies before acting on it with confidence.

```{r}
#| label: setup
source(here::here("R/utils.R"))
grade <- read_grade_data()
```

```{r}
#| label: grade-table
#| results: asis
if (nrow(grade) == 0) {
  cli::cat_line("No GRADE assessments available yet.")
} else {
  certainty_colours <- c(
    high = "#2e8a43ff",
    moderate = "#b58e2cff",
    low = "#d16d1cff",
    `very low` = "#b83744ff"
  )

  grade$group <- dplyr::coalesce(grade$domain, grade$search_topic, "other")
  grade$group <- gsub("_", " ", grade$group) |> tools::toTitleCase()

  display_cols <- intersect(
    c("outcome", "n_studies", "n_participants", "effect_estimate", "certainty"),
    names(grade)
  )
  display_names <- c(
    outcome = "Outcome",
    n_studies = "Studies",
    n_participants = "Participants",
    effect_estimate = "Effect Estimate",
    certainty = "Certainty"
  )[display_cols]

  grade$outcome <- sub("^[^:]+:\\s*", "", grade$outcome)

  for (grp in unique(grade$group)) {
    cli::cat_line(glue::glue("\n\n## {grp}\n"))
    sub <- grade[grade$group == grp, , drop = FALSE]
    df <- sub[, display_cols, drop = FALSE]
    names(df) <- display_names[display_cols]

    tbl <- display_table(df)

    if ("Certainty" %in% names(df)) {
      for (cert in names(certainty_colours)) {
        rows <- which(tolower(df$Certainty) == cert)
        if (length(rows) > 0) {
          tbl <- tbl |>
            gt::tab_style(
              style = list(
                gt::cell_text(color = certainty_colours[cert], weight = "bold")
              ),
              locations = gt::cells_body(columns = "Certainty", rows = rows)
            )
        }
      }
    }

    print(gt::as_raw_html(tbl))
  }
}
```

::: {.callout-note}
## Living Review

This page is part of a living systematic review.
GRADE assessments are updated as new studies are added and existing assessments are revised.
Certainty ratings may change in either direction as the evidence base grows.
:::
