---
title: "Outcomes & Recovery"
aliases:
  - /reviews/outcomes/
  - /reviews/recovery/
---

```{r}
#| label: setup
#| include: false
source(here::here("R/utils.R"))
source(here::here("R/meta-analysis.R"))
source(here::here("R/plots.R"))

studies <- read_study_database()
outcomes <- read_extracted_data("outcomes")
has_data <- nrow(outcomes) > 0
```

## Overview

Outcomes data tracks what happens to patients over time in the absence of specific interventions.
This is distinct from treatment trials -- here we are looking at the natural history of Long COVID and ME/CFS.
Understanding natural trajectories matters because without a clear baseline of what happens without treatment, it is impossible to judge whether an intervention actually helps.
This review synthesises longitudinal data on fatigue, physical function, quality of life, PEM, and recovery rates, stratified by PEM status where available.

```{r}
#| label: included-studies
#| results: asis
topic_studies <- studies |>
  dplyr::filter(
    !is.na(search_topic), search_topic == "outcomes",
    !is.na(llm_relevant), llm_relevant
  )
if (nrow(topic_studies) > 0) {
  cli::cat_line(glue::glue("**{nrow(topic_studies)} relevant studies** identified."))
}
```

## Outcome Domains {#sec-domains}

Studies measure patient outcomes in different ways -- some track fatigue severity, others physical function scores, others quality of life indices.
The table below shows how many findings we have in each domain.
Domains with more findings give us more confidence in pooled estimates; domains with few findings are areas where more research is needed.

```{r}
#| label: domain-summary
#| eval: !expr has_data
if (has_data && "outcome_domain" %in% names(outcomes)) {
  domain_counts <- outcomes |>
    dplyr::filter(!is.na(outcome_domain)) |>
    dplyr::count(outcome_domain, sort = TRUE)
  if (nrow(domain_counts) > 0) {
    names(domain_counts) <- c("Outcome Domain", "Findings")
    display_table(domain_counts)
  }
}
```

## Recovery Rates {#sec-recovery}

Recovery rates are among the most contested numbers in Long COVID research.
A study's reported recovery rate depends entirely on how it defines "recovery" -- return to work, symptom resolution, self-reported improvement, or something else.
The mean recovery rate below is a rough summary; the forest plot shows individual study estimates so you can see the full spread.

```{r}
#| label: recovery-summary
#| results: asis
#| eval: !expr has_data
if (has_data && all(c("recovery_n", "n_total") %in% names(outcomes))) {
  rec_df <- outcomes |>
    dplyr::filter(!is.na(recovery_n), !is.na(n_total)) |>
    dplyr::mutate(recovery_pct = recovery_n / n_total * 100)
  if (nrow(rec_df) > 0) {
    mean_rec <- round(mean(rec_df$recovery_pct, na.rm = TRUE), 1)
    cli::cat_line(glue::glue("Mean recovery rate across {nrow(rec_df)} datapoints: **{mean_rec}%**"))
  }
}
```

```{r}
#| label: recovery-data
#| include: false
rec_plot_df <- NULL
recovery_h <- 6
n_recovery_total <- 0L
if (has_data && all(c("recovery_n", "n_total") %in% names(outcomes))) {
  rec_df <- outcomes |>
    dplyr::filter(!is.na(recovery_n), !is.na(n_total)) |>
    dplyr::mutate(
      study_label = study_id,
      prevalence = recovery_n / n_total,
      ci_lower = prevalence - 1.96 * sqrt(prevalence * (1 - prevalence) / n_total),
      ci_upper = prevalence + 1.96 * sqrt(prevalence * (1 - prevalence) / n_total)
    )
  n_recovery_total <- nrow(rec_df)
  if (nrow(rec_df) > 50) {
    rec_df <- rec_df |>
      dplyr::arrange(prevalence) |>
      dplyr::slice_head(n = 50)
  }
  if (nrow(rec_df) >= 2) {
    rec_plot_df <- rec_df
    recovery_h <- forest_height(nrow(rec_df))
  }
}
```

The forest plot below displays each study's recovery estimate as a point with a 95% confidence interval.
Wide intervals indicate small samples or high uncertainty.
When studies cluster in a narrow band, the evidence is more consistent; when they scatter widely, heterogeneity in study populations or definitions is likely driving the differences.

```{r}
#| label: recovery-plot
#| eval: !expr '!is.null(rec_plot_df)'
#| fig-height: !expr recovery_h
#| fig-cap: !expr 'if (n_recovery_total > 50) paste0("Recovery rates (50 lowest of ", n_recovery_total, " datapoints)") else "Reported recovery rates by study"'
#| fig-alt: "Forest plot showing individual study recovery rate estimates as points with horizontal 95% confidence interval lines, ordered by effect size."
prevalence_plot(rec_plot_df, title = "Recovery Rates")
```

## Recovery Definitions {#sec-definitions}

How a study defines "recovery" profoundly shapes its reported recovery rate.
A study that counts "return to work" as recovery will report higher rates than one requiring full symptom resolution.
This table catalogues the definitions used across studies so you can judge which rates are comparable.

```{r}
#| label: recovery-definitions
#| eval: !expr has_data
if (has_data && "recovery_definition" %in% names(outcomes)) {
  def_counts <- outcomes |>
    dplyr::filter(!is.na(recovery_definition), recovery_definition != "") |>
    dplyr::count(recovery_definition, sort = TRUE)
  if (nrow(def_counts) > 0) {
    names(def_counts) <- c("Recovery Definition", "Studies")
    display_table(def_counts)
  }
}
```

## Outcome Trajectories {#sec-trajectories}

The trajectory plot shows how much each outcome domain changed between baseline and final follow-up.
Positive values mean improvement; negative values mean worsening.
A box that straddles the zero line suggests that some patients improved while others did not, or that improvement was modest relative to measurement variability.

```{r}
#| label: trajectory-data
#| include: false
traj_df <- NULL
traj_h <- 6
if (has_data && all(c("baseline_mean", "final_mean", "outcome_domain") %in% names(outcomes))) {
  tmp <- outcomes |>
    dplyr::filter(!is.na(baseline_mean), !is.na(final_mean), !is.na(outcome_domain)) |>
    dplyr::mutate(change = final_mean - baseline_mean)
  if (nrow(tmp) >= 2) {
    traj_df <- tmp
    traj_h <- bar_height(dplyr::n_distinct(traj_df$outcome_domain))
  }
}
```

```{r}
#| label: trajectory-plot
#| eval: !expr '!is.null(traj_df)'
#| fig-height: !expr traj_h
#| fig-cap: "Change from baseline to final follow-up by outcome domain -- boxes above the dashed zero line indicate improvement, boxes below indicate worsening."
#| fig-alt: "Horizontal box-and-jitter plot with outcome domains on the y-axis and change from baseline on the x-axis. A dashed vertical line at zero separates improvement (right) from worsening (left). Individual data points are overlaid as jittered dots."
library(ggplot2)
ggplot(traj_df, aes(x = reorder(outcome_domain, change, FUN = median), y = change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_boxplot(fill = "#0072B2", alpha = 0.4, outlier.shape = 16) +
  geom_jitter(width = 0.1, size = 2, alpha = 0.6, colour = "#0072B2") +
  coord_flip() +
  labs(
    x = NULL,
    y = "Change from baseline (final \u2212 baseline mean)",
    title = "Outcome Changes by Domain"
  ) +
  theme_review
```

## PEM Prevalence {#sec-pem}

Post-exertional malaise (PEM) is the symptom most likely to distinguish Long COVID and ME/CFS from ordinary post-viral fatigue.
When outcomes studies report PEM prevalence in their cohorts, it helps us understand whether the sample includes the patients most affected by these conditions.
A cohort with low PEM prevalence may have better outcomes simply because it contains fewer severely affected patients, not because recovery is common.

```{r}
#| label: pem-prevalence
#| results: asis
#| eval: !expr has_data
if (has_data && "pem_prevalence" %in% names(outcomes)) {
  pem_df <- outcomes |> dplyr::filter(!is.na(pem_prevalence))
  if (nrow(pem_df) > 0) {
    pooled_pem <- round(mean(pem_df$pem_prevalence, na.rm = TRUE) * 100, 1)
    cli::cat_line(glue::glue(
      "PEM prevalence reported in {nrow(pem_df)} outcomes. Mean reported prevalence: **{pooled_pem}%**."
    ))
  } else {
    cli::cat_line("No PEM prevalence estimates extracted yet.")
  }
}
```

## Condition Breakdown {#sec-condition}

Separating findings by condition lets you see whether outcome patterns differ between Long COVID and ME/CFS, or whether the two conditions follow similar trajectories.

```{r}
#| label: condition-table
#| eval: !expr has_data
if (has_data && "condition" %in% names(outcomes)) {
  cond_counts <- outcomes |>
    dplyr::filter(!is.na(condition)) |>
    dplyr::count(condition, sort = TRUE)
  names(cond_counts) <- c("Condition", "Findings")
  display_table(cond_counts)
}
```

::: {.callout-note}
## Living Review

This page is part of a living systematic review and is updated as new studies are identified, screened, and extracted.
The data, figures, and pooled estimates will change over time.

```{r}
#| label: data-status
#| results: asis
if (!has_data) {
  cli::cat_line("No extracted data yet. Run the extraction pipeline to populate this review.")
} else {
  n_studies <- dplyr::n_distinct(outcomes$study_id)
  cli::cat_line(glue::glue("{nrow(outcomes)} outcome datapoints from {n_studies} studies."))
}
```
:::
