---
title: "Outcomes & Recovery"
aliases:
  - /reviews/outcomes/
  - /reviews/recovery/
---

```{r}
#| label: setup
#| include: false
source(here::here("R/utils.R"))
source(here::here("R/meta-analysis.R"))
source(here::here("R/plots.R"))

studies <- read_study_database()
outcomes <- read_extracted_data("outcomes")
has_data <- nrow(outcomes) > 0
```

## Overview

This review synthesises longitudinal outcome and recovery data for Long COVID and ME/CFS, covering fatigue, physical function, quality of life, PEM, and recovery rates. Studies are stratified by PEM status where available.

```{r}
#| label: included-studies
#| results: asis
topic_studies <- studies |>
  dplyr::filter(
    !is.na(search_topic), search_topic == "outcomes",
    !is.na(llm_relevant), llm_relevant
  )
if (nrow(topic_studies) > 0) {
  cli::cat_line(glue::glue("**{nrow(topic_studies)} relevant studies** identified."))
}
```

## Outcome Domains {#sec-domains}

```{r}
#| label: domain-summary
#| eval: !expr has_data
if (has_data && "outcome_domain" %in% names(outcomes)) {
  domain_counts <- outcomes |>
    dplyr::filter(!is.na(outcome_domain)) |>
    dplyr::count(outcome_domain, sort = TRUE)
  if (nrow(domain_counts) > 0) {
    names(domain_counts) <- c("Outcome Domain", "Findings")
    display_table(domain_counts)
  }
}
```

## Recovery Rates {#sec-recovery}

```{r}
#| label: recovery-summary
#| results: asis
#| eval: !expr has_data
if (has_data && all(c("recovery_n", "n_total") %in% names(outcomes))) {
  rec_df <- outcomes |>
    dplyr::filter(!is.na(recovery_n), !is.na(n_total)) |>
    dplyr::mutate(recovery_pct = recovery_n / n_total * 100)
  if (nrow(rec_df) > 0) {
    mean_rec <- round(mean(rec_df$recovery_pct, na.rm = TRUE), 1)
    cli::cat_line(glue::glue("Mean recovery rate across {nrow(rec_df)} datapoints: **{mean_rec}%**"))
  }
}
```

```{r}
#| label: recovery-data
#| include: false
rec_plot_df <- NULL
recovery_h <- 6
n_recovery_total <- 0L
if (has_data && all(c("recovery_n", "n_total") %in% names(outcomes))) {
  rec_df <- outcomes |>
    dplyr::filter(!is.na(recovery_n), !is.na(n_total)) |>
    dplyr::mutate(
      study_label = study_id,
      prevalence = recovery_n / n_total,
      ci_lower = prevalence - 1.96 * sqrt(prevalence * (1 - prevalence) / n_total),
      ci_upper = prevalence + 1.96 * sqrt(prevalence * (1 - prevalence) / n_total)
    )
  n_recovery_total <- nrow(rec_df)
  if (nrow(rec_df) > 50) {
    rec_df <- rec_df |>
      dplyr::arrange(prevalence) |>
      dplyr::slice_head(n = 50)
  }
  if (nrow(rec_df) >= 2) {
    rec_plot_df <- rec_df
    recovery_h <- forest_height(nrow(rec_df))
  }
}
```

```{r}
#| label: recovery-plot
#| eval: !expr '!is.null(rec_plot_df)'
#| fig-height: !expr recovery_h
#| fig-cap: !expr 'if (n_recovery_total > 50) paste0("Recovery rates (50 lowest of ", n_recovery_total, " datapoints)") else "Reported recovery rates by study"'
prevalence_plot(rec_plot_df, title = "Recovery Rates")
```

## Recovery Definitions {#sec-definitions}

```{r}
#| label: recovery-definitions
#| eval: !expr has_data
if (has_data && "recovery_definition" %in% names(outcomes)) {
  def_counts <- outcomes |>
    dplyr::filter(!is.na(recovery_definition), recovery_definition != "") |>
    dplyr::count(recovery_definition, sort = TRUE)
  if (nrow(def_counts) > 0) {
    names(def_counts) <- c("Recovery Definition", "Studies")
    display_table(def_counts)
  }
}
```

## Outcome Trajectories {#sec-trajectories}

```{r}
#| label: trajectory-data
#| include: false
traj_df <- NULL
traj_h <- 6
if (has_data && all(c("baseline_mean", "final_mean", "outcome_domain") %in% names(outcomes))) {
  tmp <- outcomes |>
    dplyr::filter(!is.na(baseline_mean), !is.na(final_mean), !is.na(outcome_domain)) |>
    dplyr::mutate(change = final_mean - baseline_mean)
  if (nrow(tmp) >= 2) {
    traj_df <- tmp
    traj_h <- bar_height(dplyr::n_distinct(traj_df$outcome_domain))
  }
}
```

```{r}
#| label: trajectory-plot
#| eval: !expr '!is.null(traj_df)'
#| fig-height: !expr traj_h
#| fig-cap: "Outcome trajectories over follow-up (baseline vs final mean)"
library(ggplot2)
ggplot(traj_df, aes(x = reorder(outcome_domain, change, FUN = median), y = change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_boxplot(fill = "#0072B2", alpha = 0.4, outlier.shape = 16) +
  geom_jitter(width = 0.1, size = 2, alpha = 0.6, colour = "#0072B2") +
  coord_flip() +
  labs(
    x = NULL,
    y = "Change from baseline (final \u2212 baseline mean)",
    title = "Outcome Changes by Domain"
  ) +
  theme_review
```

## PEM Prevalence {#sec-pem}

```{r}
#| label: pem-prevalence
#| results: asis
#| eval: !expr has_data
if (has_data && "pem_prevalence" %in% names(outcomes)) {
  pem_df <- outcomes |> dplyr::filter(!is.na(pem_prevalence))
  if (nrow(pem_df) > 0) {
    pooled_pem <- round(mean(pem_df$pem_prevalence, na.rm = TRUE) * 100, 1)
    cli::cat_line(glue::glue(
      "PEM prevalence reported in {nrow(pem_df)} outcomes. Mean reported prevalence: **{pooled_pem}%**."
    ))
  } else {
    cli::cat_line("No PEM prevalence estimates extracted yet.")
  }
}
```

## Condition Breakdown {#sec-condition}

```{r}
#| label: condition-table
#| eval: !expr has_data
if (has_data && "condition" %in% names(outcomes)) {
  cond_counts <- outcomes |>
    dplyr::filter(!is.na(condition)) |>
    dplyr::count(condition, sort = TRUE)
  names(cond_counts) <- c("Condition", "Findings")
  display_table(cond_counts)
}
```

::: {.callout-note}
## Data Status

```{r}
#| label: data-status
#| results: asis
if (!has_data) {
  cli::cat_line("No extracted data yet. Run the extraction pipeline to populate this review.")
} else {
  n_studies <- dplyr::n_distinct(outcomes$study_id)
  cli::cat_line(glue::glue("{nrow(outcomes)} outcome datapoints from {n_studies} studies."))
}
```
:::
