---
title: "PEM Prevalence"
---

```{r}
#| label: setup
#| include: false
source("../R/utils.R")
source("../R/meta-analysis.R")
source("../R/plots.R")

studies <- read_study_database()
pem_data <- read_extracted_data("pem_prevalence")
has_data <- nrow(pem_data) > 0
```

## Overview

Post-exertional malaise (PEM) is the cardinal symptom distinguishing ME/CFS and a significant subset of Long COVID from other fatigue conditions. This section pools PEM prevalence estimates across studies and examines variation by population, assessment method, and diagnostic criteria.

## Pooled PEM Prevalence

### In Long COVID

```{r}
#| label: pem-lc
#| eval: !expr has_data
#| fig-cap: "PEM prevalence in Long COVID populations"
pem_lc <- pem_data[pem_data$condition == "long_covid", ]
if (nrow(pem_lc) >= 2) {
  fit <- run_prevalence_meta(pem_lc)
  bt <- fit$back_transform(coef(fit), fit$ci.lb, fit$ci.ub)
  cat(sprintf(
    "Pooled PEM prevalence in Long COVID: **%.1f%%** (95%% CI: %.1f%%–%.1f%%)\n\n",
    bt$prevalence * 100, bt$ci_lower * 100, bt$ci_upper * 100
  ))
  prevalence_plot(pem_lc, title = "PEM Prevalence in Long COVID")
}
```

### In ME/CFS

```{r}
#| label: pem-mecfs
#| eval: !expr has_data
#| fig-cap: "PEM prevalence in ME/CFS populations"
pem_mecfs <- pem_data[pem_data$condition == "mecfs", ]
if (nrow(pem_mecfs) >= 2) {
  fit_mecfs <- run_prevalence_meta(pem_mecfs)
  prevalence_plot(pem_mecfs, title = "PEM Prevalence in ME/CFS")
}
```

## By Assessment Method

```{r}
#| label: pem-by-method
#| eval: !expr has_data
#| fig-cap: "PEM prevalence by assessment method"
if (has_data && any(!is.na(pem_data$assessment_method))) {
  methods <- unique(pem_data$assessment_method[!is.na(pem_data$assessment_method)])
  for (m in methods) {
    m_data <- pem_data[pem_data$assessment_method == m, ]
    if (nrow(m_data) >= 2) {
      cat(sprintf("\n### %s\n\n", m))
      fit_m <- run_prevalence_meta(m_data)
      bt <- fit_m$back_transform(coef(fit_m), fit_m$ci.lb, fit_m$ci.ub)
      cat(sprintf("Prevalence: %.1f%% (%.1f%%–%.1f%%)\n\n", bt$prevalence * 100, bt$ci_lower * 100, bt$ci_upper * 100))
    }
  }
}
```

## By Diagnostic Criteria

```{r}
#| label: pem-by-criteria
#| eval: !expr has_data
if (has_data && any(!is.na(pem_data$diagnostic_criteria))) {
  cat("*Subgroup analysis by diagnostic criteria stringency will appear once data is sufficient.*\n")
}
```

## Lay Summary {#sec-lay}

```{r}
#| label: lay-pem
#| eval: !expr has_data
cat("*How common is PEM? Visual summary for non-specialist audiences.*\n")
```

::: {.callout-note}
## Data Status

```{r}
#| label: data-status
if (!has_data) {
  cat("No extracted PEM prevalence data yet.\n")
} else {
  cat(sprintf("%d estimates from %d studies.\n", nrow(pem_data), length(unique(pem_data$study_id))))
}
```
:::
