---
title: "PEM Prevalence"
---

```{r}
#| label: setup
#| include: false
source("../R/utils.R")
source("../R/meta-analysis.R")
source("../R/plots.R")

studies <- read_study_database()
pem_data <- read_extracted_data("pem_prevalence")
has_data <- nrow(pem_data) > 0
```

## Overview

Post-exertional malaise (PEM) -- the worsening of symptoms after physical or cognitive exertion that would not have caused problems before illness -- is the single most important symptom for distinguishing ME/CFS from other fatigue conditions.
In Long COVID, PEM identifies the subset of patients whose illness most closely resembles ME/CFS, and who are most likely to be harmed by graded exercise programmes.
How common PEM actually is depends heavily on how you ask: a single screening question captures fewer cases than a validated questionnaire, and diagnostic criteria that require PEM will by definition find it in 100% of qualifying patients.
This section pools PEM prevalence estimates and breaks them down by method and criteria so you can see how those choices shape the numbers.

## Pooled PEM Prevalence

### In Long COVID

The pooled estimate below combines all Long COVID studies that reported PEM prevalence, regardless of assessment method.
The forest plot shows each study's estimate so you can judge the spread.

```{r}
#| label: pem-lc
#| eval: !expr has_data
#| fig-cap: "Forest plot of PEM prevalence across Long COVID cohorts -- each row is one study, with the pooled estimate at the bottom."
#| fig-alt: "Forest plot showing PEM prevalence estimates in Long COVID populations. Each study is a row with a point estimate and horizontal confidence interval whisker. A diamond at the bottom represents the pooled random-effects estimate."
pem_lc <- pem_data[pem_data$condition == "long_covid", ]
if (nrow(pem_lc) >= 2) {
  fit <- run_prevalence_meta(pem_lc)
  bt <- fit$back_transform(coef(fit), fit$ci.lb, fit$ci.ub)
  cat(sprintf(
    "Pooled PEM prevalence in Long COVID: **%.1f%%** (95%% CI: %.1f%%–%.1f%%)\n\n",
    bt$prevalence * 100, bt$ci_lower * 100, bt$ci_upper * 100
  ))
  prevalence_plot(pem_lc, title = "PEM Prevalence in Long COVID")
}
```

### In ME/CFS

PEM prevalence in ME/CFS cohorts depends almost entirely on which diagnostic criteria were used to define the sample.
Criteria that require PEM (such as the Canadian Consensus Criteria or the IOM criteria) will report near-universal prevalence by design.
Studies using broader criteria like Oxford or Fukuda may include patients without PEM, yielding lower prevalence estimates.

```{r}
#| label: pem-mecfs
#| eval: !expr has_data
#| fig-cap: "Forest plot of PEM prevalence across ME/CFS cohorts, reflecting variation by diagnostic criteria."
#| fig-alt: "Forest plot showing PEM prevalence estimates in ME/CFS populations. Each study is a row with a point estimate and horizontal confidence interval whisker."
pem_mecfs <- pem_data[pem_data$condition == "mecfs", ]
if (nrow(pem_mecfs) >= 2) {
  fit_mecfs <- run_prevalence_meta(pem_mecfs)
  prevalence_plot(pem_mecfs, title = "PEM Prevalence in ME/CFS")
}
```

## By Assessment Method

The way PEM is measured substantially affects reported prevalence.
A single yes/no screening question will miss patients who experience PEM but do not recognise the term.
Validated instruments like the DePaul Symptom Questionnaire (DSQ) capture more cases because they describe the experience in concrete terms.
Activity monitoring and two-day cardiopulmonary exercise testing (CPET) provide objective evidence but are rarely used in large cohorts.

```{r}
#| label: pem-by-method
#| eval: !expr has_data
#| fig-cap: "PEM prevalence estimates broken down by the assessment method used in each study."
#| fig-alt: "Text output showing pooled PEM prevalence and confidence intervals grouped by assessment method (e.g. questionnaire, clinical interview, CPET)."
if (has_data && any(!is.na(pem_data$assessment_method))) {
  methods <- unique(pem_data$assessment_method[!is.na(pem_data$assessment_method)])
  for (m in methods) {
    m_data <- pem_data[pem_data$assessment_method == m, ]
    if (nrow(m_data) >= 2) {
      cat(sprintf("\n### %s\n\n", m))
      fit_m <- run_prevalence_meta(m_data)
      bt <- fit_m$back_transform(coef(fit_m), fit_m$ci.lb, fit_m$ci.ub)
      cat(sprintf("Prevalence: %.1f%% (%.1f%%–%.1f%%)\n\n", bt$prevalence * 100, bt$ci_lower * 100, bt$ci_upper * 100))
    }
  }
}
```

## By Diagnostic Criteria

Different diagnostic criteria set different bars for who counts as a patient.
Stricter criteria that require PEM will mechanically produce higher PEM prevalence; broader criteria that treat PEM as optional will produce lower estimates.
Understanding this relationship is essential for interpreting any PEM prevalence number.

```{r}
#| label: pem-by-criteria
#| eval: !expr has_data
if (has_data && any(!is.na(pem_data$diagnostic_criteria))) {
  cat("*Subgroup analysis by diagnostic criteria stringency will appear once data is sufficient.*\n")
}
```

## Lay Summary {#sec-lay}

```{r}
#| label: lay-pem
#| eval: !expr has_data
cat("*How common is PEM? Visual summary for non-specialist audiences.*\n")
```

::: {.callout-note}
## Living Review

This page is part of a living systematic review and is updated as new studies are identified, screened, and extracted.
The data, figures, and pooled estimates will change over time.
PEM prevalence estimates are particularly sensitive to new studies that use different assessment methods.

```{r}
#| label: data-status
if (!has_data) {
  cat("No extracted PEM prevalence data yet.\n")
} else {
  cat(sprintf("%d estimates from %d studies.\n", nrow(pem_data), length(unique(pem_data$study_id))))
}
```
:::
