---
title: "Prevalence"
subtitle: "How common are Long COVID and ME/CFS symptoms across populations?"
---

```{r}
#| label: setup
#| include: false
source("../R/utils.R")
source("../R/meta-analysis.R")
source("../R/plots.R")

studies <- read_study_database()
prev_data <- read_extracted_data("prevalence")
has_data <- nrow(prev_data) > 0
```

::: {.lead}
Prevalence estimates tell us how common each symptom is among people with Long COVID or ME/CFS.
These numbers vary enormously across studies -- sometimes by a factor of five or more for the same symptom -- and the reasons matter.
Studies that recruit from hospital follow-up clinics capture sicker patients than population-based surveys.
Studies that use stricter diagnostic criteria tend to find higher symptom prevalence within their smaller, more severely affected cohorts.
The pooled estimates below attempt to account for this heterogeneity, but they should be read as ranges rather than single definitive numbers.
:::

::: {.at-a-glance}
## At a Glance {#sec-lay}

The most commonly reported symptoms in Long COVID are fatigue, cognitive difficulties, and shortness of breath.
The exact numbers vary widely between studies because of different populations and measurement methods.

```{r}
#| label: study-count-badge
#| eval: !expr has_data
#| results: asis
if (has_data) {
  cat(sprintf(
    '<div class="study-count"><span class="count-number">%d</span><span class="count-label">prevalence estimates from %d studies</span></div>\n',
    nrow(prev_data), length(unique(prev_data$study_id))
  ))
}
```

```{r}
#| label: lay-symptom-bar
#| eval: !expr has_data
#| fig-height: 6
#| fig-alt: "Horizontal bar chart showing how often each symptom is reported across Long COVID studies."
#| fig-cap: "Which symptoms are most commonly studied? Each bar shows how many studies reported on that symptom. More bars = more evidence about that symptom."
if (has_data && "symptom" %in% names(prev_data)) {
  sym_counts <- prev_data |>
    dplyr::filter(!is.na(symptom)) |>
    dplyr::count(symptom, sort = TRUE) |>
    dplyr::slice_head(n = 15) |>
    dplyr::mutate(symptom = factor(symptom, levels = rev(symptom)))
  if (nrow(sym_counts) > 0) {
    ggplot(sym_counts, aes(x = n, y = symptom)) +
      geom_col(fill = cols_review["clinical"], width = 0.7) +
      geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
      scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
      labs(x = "Number of study findings", y = NULL,
           title = "Most commonly studied symptoms",
           subtitle = "More studies means more evidence about that symptom") +
      theme_lay
  }
}
```
:::

<hr class="section-divider">

::: {.panel-tabset}

## Long COVID Symptom Prevalence

Each symptom with at least two contributing studies gets a pooled estimate using a random-effects meta-analysis of proportions.
The confidence intervals reflect both within-study sampling error and between-study heterogeneity.
Wide intervals are the norm here -- they reflect genuine variation in who gets studied and how symptoms are measured.

```{r}
#| label: lc-prevalence
#| eval: !expr has_data
#| fig-cap: "Pooled prevalence of Long COVID symptoms with 95% confidence intervals from random-effects meta-analysis."
#| fig-alt: "Text output showing pooled prevalence percentages and confidence intervals for each Long COVID symptom that has at least two contributing studies."
lc <- prev_data[prev_data$condition == "long_covid", ]

if (nrow(lc) >= 2) {
  symptoms <- unique(lc$symptom)

  for (sym in symptoms) {
    sym_data <- lc[lc$symptom == sym, ]
    if (nrow(sym_data) >= 2) {
      fit <- run_prevalence_meta(sym_data)
      bt <- fit$back_transform(coef(fit), fit$ci.lb, fit$ci.ub)
      cat(sprintf(
        "\n### %s\n\nPooled prevalence: **%.1f%%** (95%% CI: %.1f%%–%.1f%%)\n\n",
        sym, bt$prevalence * 100, bt$ci_lower * 100, bt$ci_upper * 100
      ))
    }
  }
}
```

## ME/CFS Symptom Prevalence

ME/CFS prevalence data comes from a longer research history, but sample sizes tend to be smaller and diagnostic criteria vary more widely.
The forest plot below shows individual study estimates alongside the pooled result.

```{r}
#| label: mecfs-prevalence
#| eval: !expr has_data
#| fig-cap: "Forest plot of ME/CFS symptom prevalence estimates with pooled random-effects summary."
#| fig-alt: "Forest plot showing individual study prevalence estimates for ME/CFS symptoms as points with confidence interval whiskers, ordered vertically by study."
mecfs <- prev_data[prev_data$condition == "mecfs", ]
if (nrow(mecfs) >= 2) {
  fit_mecfs <- run_prevalence_meta(mecfs)
  prevalence_plot(mecfs, title = "ME/CFS Symptom Prevalence")
}
```

:::

## By Timepoint

Symptom prevalence can change over time -- some symptoms resolve, others emerge months after the initial infection.
Tracking prevalence by months post-infection reveals which symptoms are transient and which persist.

```{r}
#| label: prevalence-time
#| eval: !expr has_data
#| fig-cap: "Symptom prevalence estimates plotted by months post-infection to reveal temporal patterns."
#| fig-alt: "Scatter plot of symptom prevalence on the y-axis against months post-infection on the x-axis, showing how prevalence changes over time."
if (has_data && any(!is.na(prev_data$timepoint_months))) {
  cat("*Prevalence estimates over time will appear here once sufficient data is extracted.*\n")
}
```

::: {.callout-note collapse="true"}
## Living Review

This page is part of a living systematic review and is updated as new studies are identified, screened, and extracted.
The data, figures, and pooled estimates will change over time.
Prevalence ranges are expected to narrow as more studies are added.

```{r}
#| label: data-status
if (!has_data) {
  cat("No extracted prevalence data yet. Run the search and extraction pipeline to populate this review.\n")
} else {
  cat(sprintf(
    "%d estimates from %d studies.\n",
    nrow(prev_data), length(unique(prev_data$study_id))
  ))
}
```
:::
