---
title: "Publication Bias: Protocol-to-Publication Pipeline"
---

```{r}
#| label: setup
#| include: false
library(arrow)
library(dplyr)
library(ggplot2)
library(knitr)
source(here::here("R/registry-search.R"))
source(here::here("R/plots.R"))

studies <- read_parquet(here::here("data/studies.parquet"))
registry_file <- here::here("data/extracted/registry_trials.parquet")
```

## Rationale

Publication bias — the tendency for studies with positive or statistically significant results to be published preferentially — is a well-documented threat to systematic review validity. Registered protocols and trial registrations that never produce published results may indicate suppressed null or negative findings. By systematically tracking the protocol-to-publication pipeline, we can estimate the extent of this bias in Long COVID and ME/CFS research.

This analysis uses two complementary approaches:

1. **ClinicalTrials.gov registry search**: Direct query of registered interventional and observational studies
2. **Protocol-publication matching**: Cross-referencing published protocols in our database against published results

## ClinicalTrials.gov Registry

```{r}
#| label: fetch-registry
#| cache: true
if (file.exists(registry_file)) {
  registry <- read_parquet(registry_file)
  message(sprintf("Loaded %d registry entries from cache", nrow(registry)))
} else {
  registry <- search_ctgov()
  registry <- classify_registry_status(registry)
  registry <- filter_relevant_trials(registry)
  registry <- match_registry_to_publications(registry, studies)
  write_parquet(registry, registry_file)
}

n_excluded <- sum(!registry$relevant, na.rm = TRUE)
registry_all <- registry
registry <- registry[registry$relevant %in% TRUE, ]
```

### Registration Overview

```{r}
#| label: registry-overview
cat(sprintf("**Total registered studies**: %d (after excluding %d irrelevant registrations: vaccine immunogenicity, acute prophylaxis, serology, and off-topic studies)\n\n", nrow(registry), n_excluded))

status_summary <- registry |>
  count(publication_status) |>
  mutate(
    label = case_when(
      publication_status == "results_posted" ~ "Results posted or linked publication",
      publication_status == "completed_no_results" ~ "Completed ≥2 years, no results",
      publication_status == "recently_completed" ~ "Recently completed (<2 years)",
      publication_status == "ongoing" ~ "Still ongoing",
      publication_status == "terminated" ~ "Withdrawn or terminated",
      TRUE ~ "Unknown"
    )
  ) |>
  arrange(desc(n))

kable(
  status_summary |> select(Status = label, N = n),
  caption = "Trial registration status breakdown"
)
```

```{r}
#| label: fig-registry-status
#| fig-cap: "Status of registered Long COVID and ME/CFS trials"
status_colours <- c(
  "results_posted" = "#2ca02c",
  "completed_no_results" = "#d62728",
  "recently_completed" = "#ff7f0e",
  "ongoing" = "#1f77b4",
  "terminated" = "#7f7f7f",
  "unknown" = "#bcbd22"
)

status_labels <- c(
  "results_posted" = "Results posted",
  "completed_no_results" = "Completed, no results (≥2y)",
  "recently_completed" = "Recently completed (<2y)",
  "ongoing" = "Ongoing",
  "terminated" = "Withdrawn/Terminated",
  "unknown" = "Unknown"
)

registry |>
  count(publication_status) |>
  mutate(label = status_labels[publication_status]) |>
  ggplot(aes(x = reorder(label, n), y = n, fill = publication_status)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = status_colours, guide = "none") +
  labs(x = NULL, y = "Number of trials") +
  theme_minimal(base_size = 13)
```

### Completed Trials Without Published Results

These are trials that completed at least 2 years ago but have not posted results on ClinicalTrials.gov and could not be matched to a publication in our database. Under FDAAA 811, applicable clinical trials are required to post results within 12 months of primary completion.

```{r}
#| label: tbl-orphan-registry
#| tbl-cap: "Registered trials completed ≥2 years ago without published results"
orphans <- registry |>
  filter(publication_status == "completed_no_results") |>
  arrange(desc(years_since_completion)) |>
  select(
    `NCT ID` = nct_id,
    Title = title,
    Status = status,
    `Completion` = completion_date,
    `Years since` = years_since_completion,
    Interventions = interventions,
    Enrollment = enrollment
  ) |>
  mutate(`Years since` = round(`Years since`, 1))

if (nrow(orphans) > 0) {
  kable(orphans)
} else {
  cat("No orphan trials identified (all completed trials have posted results or publications).")
}
```

### Intervention Types of Unpublished Trials

```{r}
#| label: fig-orphan-interventions
#| fig-cap: "Intervention types among completed trials without published results"
orphan_trials <- registry |>
  filter(publication_status == "completed_no_results")

if (nrow(orphan_trials) > 0) {
  orphan_trials |>
    tidyr::separate_rows(intervention_types, sep = "; ") |>
    filter(!is.na(intervention_types), intervention_types != "") |>
    count(intervention_types) |>
    ggplot(aes(x = reorder(intervention_types, n), y = n)) +
    geom_col(fill = "#d62728") +
    coord_flip() +
    labs(x = NULL, y = "Number of unpublished trials") +
    theme_minimal(base_size = 13)
}
```

## Protocol-Publication Matching

Independently of the registry search, we cross-reference protocols published in our database (study protocols published in journals) against published results.

```{r}
#| label: protocol-matching
protocols <- match_protocols_to_publications(studies)
n_total <- nrow(protocols)
n_published <- sum(protocols$has_published_results)
n_orphan <- n_total - n_published

cat(sprintf(
  "Of **%d published protocols** in our database, **%d (%0.0f%%)** could be matched to a published results paper, and **%d (%0.0f%%)** remain without published results.\n",
  n_total, n_published, 100 * n_published / n_total,
  n_orphan, 100 * n_orphan / n_total
))
```

```{r}
#| label: fig-protocol-status
#| fig-cap: "Protocol-to-publication pipeline"
if (n_total > 0) {
  data.frame(
    status = c("Results published", "No results found"),
    n = c(n_published, n_orphan)
  ) |>
    ggplot(aes(x = status, y = n, fill = status)) +
    geom_col(width = 0.6) +
    scale_fill_manual(values = c("Results published" = "#2ca02c", "No results found" = "#d62728"), guide = "none") +
    labs(x = NULL, y = "Number of protocols") +
    theme_minimal(base_size = 13)
}
```

### Orphan Protocols by Domain

```{r}
#| label: tbl-orphan-protocols
#| tbl-cap: "Published protocols without matching results, by search domain"
orphan_protocols <- protocols |>
  filter(!has_published_results) |>
  group_by(search_topic) |>
  summarise(
    n_orphan = n(),
    titles = paste(substr(title, 1, 80), collapse = "; "),
    .groups = "drop"
  )

if (nrow(orphan_protocols) > 0) {
  kable(
    orphan_protocols |>
      select(Domain = search_topic, `Orphan protocols` = n_orphan)
  )
}
```

### Detailed Orphan Protocol List

```{r}
#| label: tbl-orphan-detail
#| tbl-cap: "Protocols without published results"
orphan_detail <- protocols |>
  filter(!has_published_results) |>
  arrange(search_topic, year) |>
  select(
    Domain = search_topic,
    Year = year,
    Title = title,
    Journal = journal,
    DOI = doi
  ) |>
  mutate(Title = substr(Title, 1, 100))

if (nrow(orphan_detail) > 0) {
  kable(orphan_detail)
} else {
  cat("All protocols matched to published results.")
}
```

## Combined Assessment

```{r}
#| label: combined-summary
registry_orphans <- registry |>
  filter(publication_status == "completed_no_results") |>
  nrow()

cat(sprintf(
  paste(
    "### Summary\n\n",
    "- **ClinicalTrials.gov**: %d registered trials; %d completed ≥2 years without results\n",
    "- **Published protocols**: %d protocols; %d without matching results paper\n",
    "- **Combined unique orphan studies**: estimated %d–%d\n\n",
    "These findings suggest that publication bias may affect %0.0f–%0.0f%% of completed research ",
    "in the Long COVID and ME/CFS field, broadly consistent with estimates from other clinical domains ",
    "(~30–50%% of trials go unpublished; @song2010dissemination).\n"
  ),
  nrow(registry), registry_orphans,
  n_total, n_orphan,
  max(registry_orphans, n_orphan), registry_orphans + n_orphan,
  100 * registry_orphans / max(1, sum(registry$publication_status %in% c("results_posted", "completed_no_results"))),
  100 * n_orphan / max(1, n_total)
))
```

## Implications for This Review

Publication bias has specific implications depending on the research domain:

- **Treatment trials**: Unpublished null results inflate pooled effect sizes in meta-analyses. Our trim-and-fill analyses (see individual treatment reviews) provide one correction, but identifying specific missing trials strengthens the case.
- **Prevalence studies**: Less directly affected, as prevalence estimates are not typically directional. However, studies finding lower-than-expected prevalence may be less likely to be published.
- **Biomarker studies**: Negative biomarker findings are notoriously difficult to publish, potentially inflating the apparent discriminatory power of candidate biomarkers.

::: {.callout-note}
Protocol-to-publication matching is approximate — it relies on title similarity and NCT ID cross-referencing. Some protocols may have published results under substantially different titles or in journals not indexed in Europe PMC. Manual verification of orphan protocols is recommended before drawing firm conclusions.
:::
