---
title: "Recovery Trajectories"
subtitle: "How recovery unfolds over time, stratified by post-exertional malaise"
---

```{r}
#| label: setup
#| include: false
source("../R/utils.R")
source("../R/meta-analysis.R")
source("../R/plots.R")

studies <- read_study_database()
recovery <- read_extracted_data("recovery_trajectories")
has_data <- nrow(recovery) > 0
```

```{r}
#| label: data-pending
#| results: asis
#| echo: false
#| eval: !expr '!has_data'
cat("::: {.callout-note}\n")
cat("## Data not yet available\n\n")
cat("The recovery trajectories data extraction has not been completed yet. ")
cat("This review will be populated once the extraction pipeline includes the `recovery_trajectories` domain. ")
cat("Raw data may exist in `data/extracted/recovery_outcomes.csv` but has not been integrated into the extraction schema.\n")
cat(":::\n")
```

::: {.lead}
Recovery from Long COVID and ME/CFS is not a straight line from sick to well.
Many patients experience relapsing-remitting patterns -- feeling better for days or weeks, then crashing after exertion.
Others plateau at a level of partial recovery that persists for months or years.
The presence of post-exertional malaise (PEM) is one of the strongest predictors of slower or absent recovery, which is why this section stratifies results by PEM status wherever the data allows.
:::

::: {.at-a-glance}
## At a Glance {#sec-lay}

Recovery from these conditions is slow and unpredictable.
The presence of post-exertional malaise (PEM) — crashes after activity — is one of the strongest predictors of whether someone will recover.

```{r}
#| label: study-count-badge
#| results: asis
if (has_data) {
  cat(sprintf(
    '<div class="study-count"><span class="count-number">%d</span><span class="count-label">datapoints from %d studies</span></div>',
    nrow(recovery), length(unique(recovery$study_id))
  ))
}
```

```{r}
#| label: lay-recovery-waffle
#| eval: !expr has_data
#| fig-height: 5
#| fig-alt: "Waffle chart showing the approximate percentage of patients who recover, where each square represents 1%."
#| fig-cap: "How many people recover? Each coloured square represents 1% of patients. Recovery rates vary hugely between studies depending on how 'recovery' is defined and how long patients are followed."
if (has_data && "recovery_pct" %in% names(recovery)) {
  median_recovery <- round(median(recovery$recovery_pct[!is.na(recovery$recovery_pct)]))
  if (!is.na(median_recovery) && median_recovery > 0) {
    lay_waffle(median_recovery, 100,
               yes_label = "Recovered",
               no_label = "Not recovered",
               title = "Approximate recovery rate across studies",
               subtitle = paste0("Median across studies: ~", median_recovery, "% (but definitions vary widely)"))
  }
}
```
:::

<hr class="section-divider">

## Recovery Rates

### Overall

The pooled recovery rate below combines all available studies, regardless of how they defined recovery or how long they followed patients.
This number should be treated as a rough orientation rather than a precise answer.

```{r}
#| label: recovery-overall
#| eval: !expr has_data
if (has_data) {
  overall <- recovery[!is.na(recovery$recovery_pct), ]
  if (nrow(overall) >= 2) {
    fit <- run_prevalence_meta(
      overall,
      xi_col = "recovery_n",
      ni_col = "n"
    )
    bt <- fit$back_transform(coef(fit), fit$ci.lb, fit$ci.ub)
    cat(sprintf(
      "Pooled recovery rate: **%.1f%%** (95%% CI: %.1f%%–%.1f%%)\n",
      bt$prevalence * 100, bt$ci_lower * 100, bt$ci_upper * 100
    ))
  }
}
```

### By PEM Status

::: {.key-finding}
PEM is the strongest predictor of slower recovery. Patients with PEM consistently show lower recovery rates than those without, often dramatically so. If your cohort does not distinguish PEM-positive from PEM-negative patients, its recovery rate is likely an average that obscures two very different trajectories.
:::

```{r}
#| label: recovery-pem
#| eval: !expr has_data
#| fig-cap: "Recovery rates stratified by PEM status -- the gap between groups reveals how much PEM shapes prognosis."
#| fig-alt: "Text output showing pooled recovery rate percentages and confidence intervals separately for PEM-positive and PEM-negative patient groups."
if (has_data && any(!is.na(recovery$pem_status))) {
  pem_recovery <- recovery[!is.na(recovery$pem_status), ]

  for (status in unique(pem_recovery$pem_status)) {
    sub <- pem_recovery[pem_recovery$pem_status == status, ]
    if (nrow(sub) >= 2) {
      fit_sub <- run_prevalence_meta(sub, xi_col = "recovery_n", ni_col = "n")
      bt <- fit_sub$back_transform(coef(fit_sub), fit_sub$ci.lb, fit_sub$ci.ub)
      cat(sprintf(
        "\n**%s**: %.1f%% (95%% CI: %.1f%%–%.1f%%)\n",
        status, bt$prevalence * 100, bt$ci_lower * 100, bt$ci_upper * 100
      ))
    }
  }
}
```

::: {.panel-tabset}

## Recovery Over Time

The trajectory plot below shows recovery rates at different timepoints after infection, with separate trend lines for PEM-positive and PEM-negative groups where data is available.
Each point is one study's estimate at one timepoint; larger points represent larger samples.
The smoothed lines (loess fits) show the general trend, but the shaded confidence bands are wide -- treat the direction of the trend as more reliable than the exact position.

```{r}
#| label: recovery-trajectory
#| eval: !expr has_data
#| fig-cap: "Recovery rate over months post-infection, stratified by PEM status -- diverging lines indicate that PEM-positive patients recover more slowly or not at all."
#| fig-alt: "Scatter plot with months post-infection on the x-axis and recovery rate (percent) on the y-axis. Points are sized by sample size and coloured by PEM status. Smoothed loess trend lines with shaded confidence bands show the trajectory for each group."
if (has_data && any(!is.na(recovery$timepoint_months))) {
  trajectory <- recovery[!is.na(recovery$timepoint_months) & !is.na(recovery$recovery_pct), ]

  if (nrow(trajectory) >= 3) {
    ggplot(trajectory, aes(
      x = timepoint_months,
      y = recovery_pct / 100,
      colour = pem_status
    )) +
      geom_point(aes(size = n), alpha = 0.7) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
      scale_y_continuous(labels = scales::percent_format()) +
      scale_colour_manual(values = cols_review, name = "PEM Status") +
      labs(
        title = "Recovery Trajectories Over Time",
        x = "Months post-infection",
        y = "Recovery rate",
        size = "Sample size"
      ) +
      theme_review
  }
}
```

## Quality of Life Trajectories

Quality of life (QoL) scores complement binary recovery rates by capturing the grey area between full recovery and no improvement.
A patient who is not "recovered" by any formal definition may still have meaningfully improved quality of life -- or may have worsened.
The trajectory below shows QoL scores over time, again stratified by PEM status.

```{r}
#| label: qol-trajectory
#| eval: !expr has_data
#| fig-cap: "Quality of life scores over time by PEM status -- upward trends indicate improvement, flat or downward trends indicate persistent disability."
#| fig-alt: "Scatter plot with months post-infection on the x-axis and quality of life score on the y-axis. Points are sized by sample size and coloured by PEM status. Smoothed loess trend lines with shaded confidence bands show the trajectory for each group."
if (has_data && any(!is.na(recovery$qol_score))) {
  qol <- recovery[!is.na(recovery$qol_score), ]
  if (nrow(qol) >= 3) {
    ggplot(qol, aes(x = timepoint_months, y = qol_score, colour = pem_status)) +
      geom_point(aes(size = n), alpha = 0.7) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
      scale_colour_manual(values = cols_review, name = "PEM Status") +
      labs(
        title = "Quality of Life Over Time",
        x = "Months post-infection",
        y = "QoL Score"
      ) +
      theme_review
  }
}
```

:::

::: {.callout-note collapse="true"}
## Living Review

This page is part of a living systematic review and is updated as new studies are identified, screened, and extracted.
The data, figures, and pooled estimates will change over time.
Recovery trajectory estimates are particularly sensitive to new long-term follow-up studies.

```{r}
#| label: data-status
if (!has_data) {
  cat("No extracted recovery data yet.\n")
} else {
  cat(sprintf(
    "%d datapoints from %d studies.\n",
    nrow(recovery), length(unique(recovery$study_id))
  ))
}
```
:::
