---
title: "Treatment: Biomedical Interventions"
subtitle: "Evidence for pharmacological, nutritional, and procedural approaches"
aliases:
  - /reviews/biomedical/
  - /reviews/pharmacological/
---

```{r}
#| label: setup
#| include: false
library(targets)
tar_config_set(store = here::here("_targets"))
source(here::here("R/utils.R"))
source(here::here("R/meta-analysis.R"))
source(here::here("R/plots.R"))

studies <- tar_read(studies)
effects <- tar_read(bio_effects)
quality <- tar_read(quality)
has_data <- nrow(effects) > 0
```

::: {.lead}
Biomedical interventions for Long COVID and ME/CFS span a wide range — from repurposed pharmaceuticals like low-dose naltrexone and antivirals, to nutritional supplements like CoQ10 and creatine, to procedural treatments like apheresis and transcranial stimulation.
:::

The evidence is growing fast but remains fragmented.
Most studies are small, many lack placebo controls, and the interventions target different mechanisms entirely.
Pooling across such different treatments is statistically possible but clinically questionable, so we present both the overall picture and domain-specific breakdowns below.

```{r}
#| label: included-studies
#| results: asis
#| echo: false
counts <- tar_read(bio_study_counts)
cat(sprintf('<div class="study-count"><span class="count-number">%d</span><span class="count-label">studies with<br>results</span></div>', counts$n_results))
cat(sprintf('<div class="study-count"><span class="count-number">%d</span><span class="count-label">randomised<br>controlled trials</span></div>', counts$n_rct))
cat(sprintf('<div class="study-count"><span class="count-number">%d</span><span class="count-label">registered<br>protocols</span></div>\n', counts$n_protocols))
```

```{r}
#| label: forest-data
#| include: false
forest_df <- tar_read(bio_forest)
has_forest <- nrow(forest_df) >= 2
intervention_forest_h <- 6
domain_forest_h <- 6
if (has_forest && "intervention_category" %in% names(forest_df)) {
  n_int <- length(unique(forest_df$intervention_category[!is.na(forest_df$intervention_category)]))
  intervention_forest_h <- bar_height(n_int)
}
if (has_forest && "outcome_domain" %in% names(forest_df)) {
  n_dom <- length(unique(forest_df$outcome_domain[!is.na(forest_df$outcome_domain)]))
  domain_forest_h <- bar_height(n_dom)
}
```

::: {.at-a-glance}

## At a Glance {#sec-lay}

```{r}
#| label: lay-biomed-lollipop
#| eval: !expr 'has_forest && "intervention_category" %in% names(forest_df)'
#| fig-height: 8
#| fig-alt: "Lollipop chart showing how much each biomedical treatment helps, with plain-language labels."
#| fig-cap: "How much does each biomedical treatment help? Each dot represents the pooled result from all studies of that treatment. Labels describe the size of the effect in plain language."
lay_lollipop(forest_df, group_col = "intervention_category",
             title = "Biomedical treatments: how much do they help?")
```

```{r}
#| label: lay-biomed-verdict
#| eval: !expr 'has_forest && "intervention_category" %in% names(forest_df)'
#| fig-height: 8
#| fig-alt: "Traffic-light dot plot showing whether each biomedical treatment is likely helpful, uncertain, or may not help."
#| fig-cap: "Does the treatment help? Green = evidence points towards benefit. Grey = we cannot tell from the available studies."
lay_treatment_summary(forest_df, group_col = "intervention_category",
                       title = "Biomedical treatments: does the evidence say they help?")
```

:::

<hr class="section-divider">

## Effects by Intervention {#sec-effects-intervention}

Pooling LDN with nirmatrelvir with creatine into a single number would be meaningless.
The plot below shows a separate pooled estimate for each intervention category — answering the question "what does the evidence say about _this specific_ treatment?"

```{r}
#| label: intervention-forest
#| eval: !expr 'has_forest && "intervention_category" %in% names(forest_df)'
#| fig-height: !expr intervention_forest_h
#| fig-alt: "Forest plot showing pooled effect sizes for each biomedical intervention category, with diamond size proportional to the number of contributing studies"
#| fig-cap: "Pooled effect sizes by intervention. Each row is a separate random-effects meta-analysis. Larger diamonds indicate more contributing study–outcome pairs."
subgroup_forest(
  forest_df, group_col = "intervention_category",
  title = "Pooled effects by intervention"
)
```

## Effects by Outcome Domain {#sec-effects-domain}

Same evidence, different cut.
This groups by what was measured — fatigue, cognition, pain — regardless of which drug or supplement was used.
Useful for asking: "across all biomedical approaches, which symptoms are most responsive to treatment?"

```{r}
#| label: domain-forest
#| eval: !expr 'has_forest && "outcome_domain" %in% names(forest_df)'
#| fig-height: !expr domain_forest_h
#| fig-alt: "Forest plot showing pooled effect sizes for each outcome domain across biomedical interventions"
#| fig-cap: "Pooled effect sizes by outcome domain. Each row pools all biomedical interventions that measured that domain."
subgroup_forest(
  forest_df, group_col = "outcome_domain",
  title = "Pooled effects by outcome domain"
)
```

## Prevention: Reducing Risk of Developing Long COVID {#sec-prevention}

::: {.key-finding}
Several interventions studied in the biomedical literature are not treatments for established Long COVID or ME/CFS — they are **prevention strategies** aimed at reducing the risk of developing Long COVID after SARS-CoV-2 infection.
These include vaccination before infection, antiviral treatment during acute COVID, and SSRI use during acute illness.
Their effect sizes are reported as odds ratios, hazard ratios, or relative risks (where values below 1.0 indicate a protective effect), and cannot be pooled with the standardised mean differences used in the treatment forest plots above.
:::

```{r}
#| label: prevention-data
#| include: false
prevention_df <- tar_read(bio_prevention)
has_prevention <- nrow(prevention_df) > 0
prevention_h <- if (has_prevention) min(20, max(5, nrow(prevention_df) * 0.4 + 2)) else 6
```

```{r}
#| label: prevention-forest
#| eval: !expr has_prevention
#| fig-height: !expr prevention_h
#| fig-alt: "Forest plot of prevention interventions showing odds ratios, hazard ratios, and relative risks on a log scale. Values below 1 indicate the intervention reduces risk of Long COVID."
#| fig-cap: "Prevention interventions: risk ratios on a log scale. Values left of the dashed line (ratio < 1) indicate the intervention reduces the risk of developing Long COVID. Point size is proportional to sample size where available."
ggplot(
  prevention_df,
  aes(
    y = reorder(study_label, -effect_size),
    x = effect_size,
    xmin = ci_lower,
    xmax = ci_upper,
    colour = intervention_category
  )
) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(height = 0.2) +
  geom_point(size = 2.5) +
  scale_x_log10() +
  labs(
    x = "Risk ratio (log scale)",
    y = NULL,
    colour = "Intervention"
  ) +
  theme_review +
  theme(legend.position = "bottom")
```

```{r}
#| label: prevention-table
#| eval: !expr has_prevention
prev_summary <- prevention_df |>
  dplyr::group_by(intervention_category) |>
  dplyr::summarise(
    k = dplyr::n(),
    `Median ratio` = round(median(effect_size), 2),
    `Range` = paste0(round(min(effect_size), 2), "–", round(max(effect_size), 2)),
    .groups = "drop"
  ) |>
  dplyr::arrange(`Median ratio`)
names(prev_summary)[1] <- "Intervention"
display_table(prev_summary)
```

## Interventions & Outcomes {.tabset}

::: {.panel-tabset}

### By Intervention

The table below groups specific drugs, supplements, and procedures into broader categories.
Low-dose naltrexone and nirmatrelvir/ritonavir have the most reported outcomes, though many of those come from observational rather than randomised studies.

```{r}
#| label: intervention-table
#| eval: !expr has_data
if (has_data && "intervention_category" %in% names(effects)) {
  int_counts <- effects |>
    dplyr::filter(!is.na(intervention_category)) |>
    dplyr::count(intervention_category, sort = TRUE)
  if (nrow(int_counts) > 0) {
    names(int_counts) <- c("Intervention", "Effect Sizes")
    display_table(int_counts)
  }
}
```

### By Outcome Measure

As with psychological interventions, outcome measurement varies widely.
The table below maps each instrument to its domain — making it easier to spot that, for example, three different fatigue scales are measuring the same underlying construct.

```{r}
#| label: outcome-table
#| eval: !expr has_data
if (has_data && "outcome_measure" %in% names(effects)) {
  out_counts <- effects |>
    dplyr::filter(!is.na(outcome_measure)) |>
    dplyr::mutate(
      domain = dplyr::coalesce(outcome_domain, "other")
    ) |>
    dplyr::count(domain, outcome_measure, sort = TRUE)
  if (nrow(out_counts) > 0) {
    names(out_counts) <- c("Domain", "Outcome Measure", "N")
    display_table(out_counts)
  }
}
```

### By Domain

Grouping by outcome domain cuts through the instrument heterogeneity.
The bar chart shows where the evidence concentrates; the summary table pools effect sizes within each domain.

High I^2^ values indicate substantial heterogeneity within a domain — expected, given how different the interventions are.

```{r}
#| label: domain-bar-data
#| include: false
domain_df <- NULL
domain_h <- 4
if (has_data && "outcome_domain" %in% names(effects)) {
  domain_df <- effects |>
    dplyr::filter(!is.na(outcome_domain)) |>
    dplyr::count(outcome_domain, sort = TRUE) |>
    dplyr::mutate(outcome_domain = factor(outcome_domain, levels = rev(outcome_domain)))
  domain_h <- bar_height(nrow(domain_df))
}
```

```{r}
#| label: domain-bar
#| eval: !expr '!is.null(domain_df) && nrow(domain_df) > 0'
#| fig-height: !expr domain_h
#| fig-alt: "Horizontal bar chart showing the number of extracted effect sizes per outcome domain for biomedical interventions"
#| fig-cap: "Distribution of effect sizes across outcome domains. Fatigue dominates, reflecting both clinical priority and measurement availability."
ggplot(domain_df, aes(x = n, y = outcome_domain)) +
  geom_col(fill = cols_review["clinical"], width = 0.7) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Effect sizes", y = NULL) +
  theme_review
```

```{r}
#| label: domain-forest-data
#| include: false
domain_fits <- list()
domain_forest_h <- 6
if (has_data && all(c("mean_intervention", "sd_intervention", "n_intervention",
                      "mean_comparator", "sd_comparator", "n_comparator",
                      "outcome_domain") %in% names(effects))) {
  for (dom in unique(effects$outcome_domain[!is.na(effects$outcome_domain)])) {
    dom_data <- effects |>
      dplyr::filter(
        outcome_domain == dom,
        !is.na(mean_intervention), !is.na(mean_comparator),
        !is.na(sd_intervention), !is.na(sd_comparator),
        !is.na(n_intervention), !is.na(n_comparator),
        n_intervention > 0, n_comparator > 0,
        sd_intervention > 0, sd_comparator > 0
      ) |>
      dplyr::rename(
        mean_treatment = mean_intervention, sd_treatment = sd_intervention,
        n_treatment = n_intervention, mean_control = mean_comparator,
        sd_control = sd_comparator, n_control = n_comparator
      ) |>
      dplyr::mutate(study_label = paste0(study_id, ": ", outcome_measure))
    if (nrow(dom_data) >= 2) {
      es <- compute_effect_sizes(dom_data, outcome_type = "continuous")
      domain_fits[[dom]] <- run_meta(es)
    }
  }
  if (length(domain_fits) > 0) {
    total_k <- sum(vapply(domain_fits, \(f) f$k, integer(1)))
    domain_forest_h <- forest_height(total_k + length(domain_fits))
  }
}
```

```{r}
#| label: domain-summary-table
#| eval: !expr 'length(domain_fits) > 0'
domain_summary <- purrr::imap_dfr(domain_fits, \(fit, dom) {
  data.frame(
    Domain = dom,
    k = fit$k,
    SMD = round(coef(fit), 2),
    CI = paste0("[", round(fit$ci.lb, 2), ", ", round(fit$ci.ub, 2), "]"),
    p = formatC(fit$pval, format = "g", digits = 3),
    I2 = paste0(round(fit$I2, 1), "%")
  )
}) |> dplyr::arrange(SMD)
display_table(domain_summary)
```

:::

## Dosing & Protocol {#sec-dose}

Dosing information matters for biomedical interventions in a way it does not for CBT.
The table below lists each study's reported dose or protocol, grouped by intervention category.

```{r}
#| label: dose-table
#| eval: !expr has_data
if (has_data && "dose" %in% names(effects)) {
  dose_df <- effects |>
    dplyr::filter(!is.na(dose), !is.na(intervention_category)) |>
    dplyr::select(study_id, intervention_category, dose) |>
    dplyr::distinct()
  if (nrow(dose_df) > 0) {
    names(dose_df) <- c("Study", "Intervention", "Dose / Protocol")
    display_table(dose_df)
  }
}
```

## PEM Assessment {#sec-pem}

::: {.key-finding}
Post-exertional malaise (PEM) is the hallmark symptom of ME/CFS and a defining feature of Long COVID for many patients.
Whether a biomedical trial assessed PEM matters less for safety than in exercise trials, but it is still relevant — some pharmacological interventions aim specifically to reduce PEM, and knowing whether they measured it tells us whether they could detect their own target outcome.
:::

```{r}
#| label: pem-status
#| results: asis
#| eval: !expr has_data
if (has_data && "pem_assessed" %in% names(effects)) {
  n_pem <- sum(effects$pem_assessed, na.rm = TRUE)
  pct <- round(n_pem / nrow(effects) * 100)
  cli::cat_line(glue::glue("PEM was assessed in **{n_pem}/{nrow(effects)}** ({pct}%) reported outcomes."))
}
```

## Risk of Bias {#sec-rob}

Risk of bias ratings follow the Cochrane ROB2 tool for RCTs and ROBINS-I for observational studies.

```{r}
#| label: rob-summary
#| eval: !expr nrow(quality) > 0
biomed_quality <- quality |>
  dplyr::filter(!is.na(search_topic), search_topic == "treatment_biomedical", domain_name == "overall")
if (nrow(biomed_quality) > 0) {
  rob_counts <- biomed_quality |>
    dplyr::count(judgement)
  names(rob_counts) <- c("Overall RoB", "Studies")
  display_table(rob_counts)
}
```

## Adverse Events {#sec-safety}

Adverse event reporting is inconsistent.
Some studies report detailed safety data; many say nothing at all.
Absence of adverse event reporting does not mean absence of adverse events.

```{r}
#| label: adverse-events
#| results: asis
#| eval: !expr has_data
if (has_data && "adverse_events_reported" %in% names(effects)) {
  n_ae <- sum(effects$adverse_events_reported, na.rm = TRUE)
  pct <- if (nrow(effects) > 0) round(n_ae / nrow(effects) * 100) else 0
  cli::cat_line(glue::glue("Adverse events reported in **{n_ae}/{nrow(effects)}** ({pct}%) outcomes."))
  if (n_ae > 0 && "adverse_events_detail" %in% names(effects)) {
    details <- effects$adverse_events_detail[!is.na(effects$adverse_events_detail) & effects$adverse_events_detail != ""]
    if (length(details) > 0) {
      cli::cat_line("\n**Reported adverse events:**\n")
      purrr::walk(unique(details), \(d) cli::cat_bullet(d))
    }
  }
}
```

::: {.callout-note collapse="true"}
## Data Status

```{r}
#| label: data-status
#| results: asis
if (!has_data) {
  cli::cat_line("No extracted data yet. Run the extraction pipeline to populate this review.")
} else {
  n_studies <- dplyr::n_distinct(effects$study_id)
  cli::cat_line(glue::glue("{nrow(effects)} effect sizes from {n_studies} studies."))
}
```

This is a living review.
Data is extracted automatically from published studies and updated as new evidence becomes available.
:::
