---
title: "Psychological vs Biomedical Treatments"
aliases:
  - /reviews/comparison/
---

```{r}
#| label: setup
#| include: false
source(here::here("R/utils.R"))
source(here::here("R/meta-analysis.R"))
source(here::here("R/plots.R"))

studies <- read_study_database()
cbt <- read_extracted_data("treatment_cbt_psych")
biomed <- read_extracted_data("treatment_biomedical")
quality <- read_quality_ratings()

has_cbt <- nrow(cbt) > 0
has_biomed <- nrow(biomed) > 0
has_both <- has_cbt && has_biomed
```

## Overview

This page directly compares the published research on **psychological/behavioural interventions** (CBT, GET, self-management) with **biomedical interventions** (pharmacological, immunological, device-based) for Long COVID and ME/CFS.

The comparison covers:

- Volume and growth of published studies over time
- Study designs used in each paradigm
- Reported effect sizes and outcome measures
- Risk of bias and evidence quality

```{r}
#| label: study-counts
#| results: asis
treatment_studies <- studies |>
  dplyr::filter(
    search_topic %in% c("treatment_cbt_psych", "treatment_biomedical"),
    !is.na(included), included
  )

if (nrow(treatment_studies) > 0) {
  counts <- treatment_studies |>
    dplyr::count(search_topic)
  n_psych <- counts$n[counts$search_topic == "treatment_cbt_psych"] %||% 0L
  n_bio <- counts$n[counts$search_topic == "treatment_biomedical"] %||% 0L
  cli::cat_line(glue::glue("**{n_psych} psychological/behavioural** and **{n_bio} biomedical** studies currently included."))
}
```

## Publication Trends {#sec-trends}

```{r}
#| label: publication-timeline
#| fig-cap: "Number of studies published per year by treatment paradigm"
all_treatment <- studies |>
  dplyr::filter(
    search_topic %in% c("treatment_cbt_psych", "treatment_biomedical"),
    !is.na(llm_relevant), llm_relevant
  ) |>
  dplyr::mutate(paradigm = dplyr::if_else(
    search_topic == "treatment_cbt_psych",
    "Psychological / Behavioural",
    "Biomedical"
  ))

if (nrow(all_treatment) > 0) {
  pub_counts <- all_treatment |>
    dplyr::count(Year = as.integer(year), Paradigm = paradigm)

  library(ggplot2)
  ggplot(pub_counts, aes(x = Year, y = n, colour = Paradigm)) +
    geom_point() +
    geom_line() +
    scale_colour_manual(values = c(
      "Psychological / Behavioural" = "#E69F00",
      "Biomedical" = "#0072B2"
    )) +
    labs(y = "Studies", fill = NULL) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "top")
}
```

## Study Design Comparison {#sec-designs}

```{r}
#| label: design-data
#| include: false
design_counts <- NULL
design_h <- 6
if (nrow(all_treatment) > 0) {
  design_col <- if ("llm_study_type" %in% names(all_treatment)) "llm_study_type" else "study_type"
  design_counts <- all_treatment |>
    dplyr::count(Design = .data[[design_col]], Paradigm = paradigm) |>
    dplyr::filter(n > 0)
  design_h <- bar_height(dplyr::n_distinct(design_counts$Design))
}
```

```{r}
#| label: design-comparison
#| eval: !expr '!is.null(design_counts)'
#| fig-height: !expr design_h
#| fig-cap: "Study designs used in psychological vs biomedical treatment research"
ggplot(design_counts, aes(x = reorder(Design, n), y = n, fill = Paradigm)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c(
    "Psychological / Behavioural" = "#E69F00",
    "Biomedical" = "#0072B2"
  )) +
  labs(x = NULL, y = "Studies", fill = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

## Effect Size Comparison {#sec-effects}

```{r}
#| label: effect-data
#| include: false
forest_fit <- NULL
forest_h <- 6
fit_cbt_only <- NULL
cbt_h <- 6
fit_bio_only <- NULL
bio_h <- 6

if (has_both) {
  cbt$paradigm <- "Psychological / Behavioural"
  biomed$paradigm <- "Biomedical"
  shared_cols <- intersect(names(cbt), names(biomed))
  combined <- dplyr::bind_rows(
    dplyr::select(cbt, dplyr::all_of(shared_cols)),
    dplyr::select(biomed, dplyr::all_of(shared_cols))
  )

  required <- c("mean_intervention", "sd_intervention", "n_intervention",
                 "mean_comparator", "sd_comparator", "n_comparator")
  if (all(required %in% names(combined))) {
    complete <- combined |>
      dplyr::filter(
        !is.na(mean_intervention), !is.na(mean_comparator),
        !is.na(sd_intervention), !is.na(sd_comparator),
        !is.na(n_intervention), !is.na(n_comparator),
        n_intervention > 0, n_comparator > 0,
        sd_intervention > 0, sd_comparator > 0
      )
    if (nrow(complete) >= 2) {
      complete <- complete |>
        dplyr::rename(
          mean_treatment = mean_intervention,
          sd_treatment   = sd_intervention,
          n_treatment    = n_intervention,
          mean_control   = mean_comparator,
          sd_control     = sd_comparator,
          n_control      = n_comparator
        ) |>
        dplyr::mutate(study_label = study_id)
      es <- compute_effect_sizes(complete, outcome_type = "continuous")
      forest_fit <- run_meta(es)
      forest_h <- forest_height(forest_fit$k)
    }
  }
} else if (has_cbt) {
  es_cbt <- compute_effect_sizes(cbt, outcome_type = "continuous")
  if (nrow(es_cbt) >= 2) {
    fit_cbt_only <- run_meta(es_cbt)
    cbt_h <- forest_height(fit_cbt_only$k)
  }
} else if (has_biomed) {
  es_bio <- compute_effect_sizes(biomed, outcome_type = "continuous")
  if (nrow(es_bio) >= 2) {
    fit_bio_only <- run_meta(es_bio)
    bio_h <- forest_height(fit_bio_only$k)
  }
}
```

```{r}
#| label: effect-comparison
#| eval: !expr '!is.null(forest_fit)'
#| fig-height: !expr forest_h
#| fig-cap: "Pooled effect sizes (Hedges' g) by treatment paradigm"
forest_gg(forest_fit, title = "Treatment Effects by Paradigm")
```

```{r}
#| label: effect-cbt-only
#| eval: !expr '!is.null(fit_cbt_only)'
#| fig-height: !expr cbt_h
#| results: asis
cli::cat_line("Only CBT/psychological data available so far \u2014 biomedical comparison pending.")
forest_gg(fit_cbt_only, title = "Psychological / Behavioural Interventions")
```

```{r}
#| label: effect-biomed-only
#| eval: !expr '!is.null(fit_bio_only)'
#| fig-height: !expr bio_h
#| results: asis
cli::cat_line("Only biomedical data available so far \u2014 psychological comparison pending.")
forest_gg(fit_bio_only, title = "Biomedical Interventions")
```

## Outcome Measures Used {#sec-outcomes}

```{r}
#| label: outcome-data
#| include: false
outcome_counts <- NULL
outcome_h <- 6
n_outcome_total <- 0L
if (has_both && "outcome_measure" %in% names(combined)) {
  outcome_counts <- combined |>
    dplyr::count(Measure = outcome_measure, Paradigm = paradigm) |>
    dplyr::filter(n > 0)
  n_outcome_total <- dplyr::n_distinct(outcome_counts$Measure)
  top_measures <- outcome_counts |>
    dplyr::group_by(Measure) |>
    dplyr::summarise(total = sum(n), .groups = "drop") |>
    dplyr::slice_max(total, n = 30) |>
    dplyr::pull(Measure)
  outcome_counts <- outcome_counts |>
    dplyr::filter(Measure %in% top_measures)
  outcome_h <- bar_height(dplyr::n_distinct(outcome_counts$Measure))
}
```

```{r}
#| label: outcome-measures
#| eval: !expr '!is.null(outcome_counts)'
#| fig-height: !expr outcome_h
#| fig-cap: !expr 'if (n_outcome_total > 30) paste0("Top 30 of ", n_outcome_total, " outcome measures by frequency") else "Outcome measures used in psychological vs biomedical research"'
ggplot(outcome_counts, aes(x = reorder(Measure, n), y = n, fill = Paradigm)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c(
    "Psychological / Behavioural" = "#E69F00",
    "Biomedical" = "#0072B2"
  )) +
  labs(x = NULL, y = "Effect sizes", fill = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

## Evidence Quality {#sec-quality}

```{r}
#| label: quality-comparison
#| eval: !expr nrow(quality) > 0
treatment_quality <- quality |>
  dplyr::filter(
    search_topic %in% c("treatment_cbt_psych", "treatment_biomedical"),
    domain_name == "overall"
  )

if (nrow(treatment_quality) > 0) {
  treatment_quality <- treatment_quality |>
    dplyr::mutate(paradigm = dplyr::if_else(
      search_topic == "treatment_cbt_psych",
      "Psychological / Behavioural",
      "Biomedical"
    ))

  rob_summary <- treatment_quality |>
    dplyr::count(Paradigm = paradigm, `Overall RoB` = judgement) |>
    dplyr::filter(n > 0)
  names(rob_summary) <- c("Paradigm", "Overall RoB", "Studies")
  display_table(rob_summary)
}
```

## Summary Table {#sec-summary}

```{r}
#| label: summary-table
if (nrow(all_treatment) > 0) {
  design_col <- if ("llm_study_type" %in% names(all_treatment)) "llm_study_type" else "study_type"
  summary_df <- all_treatment |>
    dplyr::mutate(is_rct = grepl("RCT|random", .data[[design_col]], ignore.case = TRUE)) |>
    dplyr::group_by(paradigm) |>
    dplyr::summarise(
      Studies = dplyr::n(),
      RCTs = sum(is_rct, na.rm = TRUE),
      `Year range` = paste(range(year, na.rm = TRUE), collapse = "–"),
      .groups = "drop"
    ) |>
    dplyr::rename(Paradigm = paradigm)
  display_table(summary_df)
}
```

::: {.callout-note}
## Data Status

```{r}
#| label: data-status
if (!has_cbt && !has_biomed) {
  cli::cat_line("No extracted treatment data yet. Run the search and extraction pipeline to populate this comparison.")
} else {
  n_cbt <- if (has_cbt) nrow(cbt) else 0L
  n_bio <- if (has_biomed) nrow(biomed) else 0L
  cli::cat_line(glue::glue("{n_cbt} psychological and {n_bio} biomedical effect sizes extracted."))
}
```
:::
