---
title: "Psychological vs Biomedical Treatments"
aliases:
  - /reviews/comparison/
---

```{r}
#| label: setup
#| include: false
source(here::here("R/utils.R"))
source(here::here("R/meta-analysis.R"))
source(here::here("R/plots.R"))

studies <- read_study_database()
cbt <- read_extracted_data("treatment_cbt_psych")
biomed <- read_extracted_data("treatment_biomedical")
quality <- read_quality_ratings()

has_cbt <- nrow(cbt) > 0
has_biomed <- nrow(biomed) > 0
has_both <- has_cbt && has_biomed
```

## Overview

Psychological and biomedical treatments for Long COVID and ME/CFS come from fundamentally different research traditions.
Psychological interventions -- CBT, graded exercise, self-management programmes -- typically assume that illness perpetuation involves cognitive or behavioural factors.
Biomedical interventions -- pharmacological agents, immunomodulators, device-based therapies -- target underlying physiological mechanisms directly.
These different starting assumptions shape which outcomes researchers measure, which study designs they favour, and how they interpret results.

This page puts those two traditions side by side so you can see where they converge and where they diverge.

The comparison covers:

- Volume and growth of published studies over time
- Study designs used in each paradigm
- Reported effect sizes and outcome measures
- Risk of bias and evidence quality

```{r}
#| label: study-counts
#| results: asis
treatment_studies <- studies |>
  dplyr::filter(
    search_topic %in% c("treatment_cbt_psych", "treatment_biomedical"),
    !is.na(included), included
  )

if (nrow(treatment_studies) > 0) {
  counts <- treatment_studies |>
    dplyr::count(search_topic)
  n_psych <- counts$n[counts$search_topic == "treatment_cbt_psych"] %||% 0L
  n_bio <- counts$n[counts$search_topic == "treatment_biomedical"] %||% 0L
  cli::cat_line(glue::glue("**{n_psych} psychological/behavioural** and **{n_bio} biomedical** studies currently included."))
}
```

## Publication Trends {#sec-trends}

Publication timelines reveal which paradigm has attracted more research attention and when.
A surge in biomedical publications after 2020 would suggest that Long COVID shifted funding priorities toward mechanistic research.
Conversely, a steady psychological output alongside that surge would indicate the two traditions are growing in parallel rather than one replacing the other.

```{r}
#| label: publication-timeline
#| fig-alt: "Line chart showing number of published studies per year, with separate lines for psychological/behavioural and biomedical treatment paradigms."
#| fig-cap: "Number of studies published per year by treatment paradigm"
all_treatment <- studies |>
  dplyr::filter(
    search_topic %in% c("treatment_cbt_psych", "treatment_biomedical"),
    !is.na(llm_relevant), llm_relevant
  ) |>
  dplyr::mutate(paradigm = dplyr::if_else(
    search_topic == "treatment_cbt_psych",
    "Psychological / Behavioural",
    "Biomedical"
  ))

if (nrow(all_treatment) > 0) {
  pub_counts <- all_treatment |>
    dplyr::count(Year = as.integer(year), Paradigm = paradigm)

  library(ggplot2)
  ggplot(pub_counts, aes(x = Year, y = n, colour = Paradigm)) +
    geom_point() +
    geom_line() +
    scale_colour_manual(values = c(
      "Psychological / Behavioural" = "#E69F00",
      "Biomedical" = "#0072B2"
    )) +
    labs(y = "Studies", fill = NULL) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "top")
}
```

## Study Design Comparison {#sec-designs}

Study design choices matter because they determine how much confidence we can place in the results.
Randomised controlled trials provide the strongest causal evidence, but they are harder to run for some interventions than others.
Biomedical research often relies on small pilot trials or case series early in development, while psychological interventions have a longer history of large-scale RCTs (particularly in the ME/CFS literature).
Comparing the design distributions helps explain why evidence quality may differ between paradigms.

```{r}
#| label: design-data
#| include: false
design_counts <- NULL
design_h <- 6
if (nrow(all_treatment) > 0) {
  design_col <- if ("llm_study_type" %in% names(all_treatment)) "llm_study_type" else "study_type"
  design_counts <- all_treatment |>
    dplyr::count(Design = .data[[design_col]], Paradigm = paradigm) |>
    dplyr::filter(n > 0)
  design_h <- bar_height(dplyr::n_distinct(design_counts$Design))
}
```

```{r}
#| label: design-comparison
#| eval: !expr '!is.null(design_counts)'
#| fig-height: !expr design_h
#| fig-alt: "Horizontal bar chart comparing the number of studies using each study design, grouped by psychological/behavioural vs biomedical paradigm."
#| fig-cap: "Study designs used in psychological vs biomedical treatment research"
ggplot(design_counts, aes(x = reorder(Design, n), y = n, fill = Paradigm)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c(
    "Psychological / Behavioural" = "#E69F00",
    "Biomedical" = "#0072B2"
  )) +
  labs(x = NULL, y = "Studies", fill = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

## Effect Size Comparison {#sec-effects}

The forest plot below pools effect sizes from both paradigms on a single scale (Hedges' g).
This is informative but requires a caveat: combining psychological and biomedical trials mixes very different interventions, populations, and control conditions.
A small overall effect from one paradigm does not mean every intervention in that paradigm is weak -- it may reflect wider variation in what "treatment" and "control" mean.
Read the individual study labels to understand what each point estimate actually represents.

```{r}
#| label: effect-data
#| include: false
forest_fit <- NULL
forest_h <- 6
fit_cbt_only <- NULL
cbt_h <- 6
fit_bio_only <- NULL
bio_h <- 6

if (has_both) {
  cbt$paradigm <- "Psychological / Behavioural"
  biomed$paradigm <- "Biomedical"
  shared_cols <- intersect(names(cbt), names(biomed))
  combined <- dplyr::bind_rows(
    dplyr::select(cbt, dplyr::all_of(shared_cols)),
    dplyr::select(biomed, dplyr::all_of(shared_cols))
  )

  required <- c("mean_intervention", "sd_intervention", "n_intervention",
                 "mean_comparator", "sd_comparator", "n_comparator")
  if (all(required %in% names(combined))) {
    complete <- combined |>
      dplyr::filter(
        !is.na(mean_intervention), !is.na(mean_comparator),
        !is.na(sd_intervention), !is.na(sd_comparator),
        !is.na(n_intervention), !is.na(n_comparator),
        n_intervention > 0, n_comparator > 0,
        sd_intervention > 0, sd_comparator > 0
      )
    if (nrow(complete) >= 2) {
      complete <- complete |>
        dplyr::rename(
          mean_treatment = mean_intervention,
          sd_treatment   = sd_intervention,
          n_treatment    = n_intervention,
          mean_control   = mean_comparator,
          sd_control     = sd_comparator,
          n_control      = n_comparator
        ) |>
        dplyr::mutate(study_label = study_id)
      es <- compute_effect_sizes(complete, outcome_type = "continuous")
      forest_fit <- run_meta(es)
      forest_h <- forest_height(forest_fit$k)
    }
  }
} else if (has_cbt) {
  es_cbt <- compute_effect_sizes(cbt, outcome_type = "continuous")
  if (nrow(es_cbt) >= 2) {
    fit_cbt_only <- run_meta(es_cbt)
    cbt_h <- forest_height(fit_cbt_only$k)
  }
} else if (has_biomed) {
  es_bio <- compute_effect_sizes(biomed, outcome_type = "continuous")
  if (nrow(es_bio) >= 2) {
    fit_bio_only <- run_meta(es_bio)
    bio_h <- forest_height(fit_bio_only$k)
  }
}
```

```{r}
#| label: effect-comparison
#| eval: !expr '!is.null(forest_fit)'
#| fig-height: !expr forest_h
#| fig-alt: "Forest plot displaying Hedges' g effect sizes for individual studies, grouped by psychological/behavioural and biomedical treatment paradigm, with pooled diamond estimates."
#| fig-cap: "Pooled effect sizes (Hedges' g) by treatment paradigm"
forest_gg(forest_fit, title = "Treatment Effects by Paradigm")
```

```{r}
#| label: effect-cbt-only
#| eval: !expr '!is.null(fit_cbt_only)'
#| fig-height: !expr cbt_h
#| fig-alt: "Forest plot of Hedges' g effect sizes for psychological and behavioural intervention studies only."
#| fig-cap: "Pooled effect sizes for psychological/behavioural interventions"
#| results: asis
cli::cat_line("Only CBT/psychological data available so far \u2014 biomedical comparison pending.")
forest_gg(fit_cbt_only, title = "Psychological / Behavioural Interventions")
```

```{r}
#| label: effect-biomed-only
#| eval: !expr '!is.null(fit_bio_only)'
#| fig-height: !expr bio_h
#| fig-alt: "Forest plot of Hedges' g effect sizes for biomedical intervention studies only."
#| fig-cap: "Pooled effect sizes for biomedical interventions"
#| results: asis
cli::cat_line("Only biomedical data available so far \u2014 psychological comparison pending.")
forest_gg(fit_bio_only, title = "Biomedical Interventions")
```

## Outcome Measures Used {#sec-outcomes}

The outcome measures each paradigm chooses reveal its assumptions about what matters.
Psychological trials tend to favour self-report questionnaires for fatigue, mood, and functional capacity.
Biomedical trials more often include laboratory markers, objective physiological tests, or clinician-rated scales.
Where the same measure appears in both paradigms, direct comparison becomes more meaningful; where measures diverge, apparent differences in effectiveness may partly reflect measurement culture rather than true treatment differences.

```{r}
#| label: outcome-data
#| include: false
outcome_counts <- NULL
outcome_h <- 6
n_outcome_total <- 0L
if (has_both && "outcome_measure" %in% names(combined)) {
  outcome_counts <- combined |>
    dplyr::count(Measure = outcome_measure, Paradigm = paradigm) |>
    dplyr::filter(n > 0)
  n_outcome_total <- dplyr::n_distinct(outcome_counts$Measure)
  top_measures <- outcome_counts |>
    dplyr::group_by(Measure) |>
    dplyr::summarise(total = sum(n), .groups = "drop") |>
    dplyr::slice_max(total, n = 30) |>
    dplyr::pull(Measure)
  outcome_counts <- outcome_counts |>
    dplyr::filter(Measure %in% top_measures)
  outcome_h <- bar_height(dplyr::n_distinct(outcome_counts$Measure))
}
```

```{r}
#| label: outcome-measures
#| eval: !expr '!is.null(outcome_counts)'
#| fig-height: !expr outcome_h
#| fig-alt: "Horizontal bar chart showing the most frequently used outcome measures, split by psychological/behavioural and biomedical paradigm."
#| fig-cap: !expr 'if (n_outcome_total > 30) paste0("Top 30 of ", n_outcome_total, " outcome measures by frequency") else "Outcome measures used in psychological vs biomedical research"'
ggplot(outcome_counts, aes(x = reorder(Measure, n), y = n, fill = Paradigm)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c(
    "Psychological / Behavioural" = "#E69F00",
    "Biomedical" = "#0072B2"
  )) +
  labs(x = NULL, y = "Effect sizes", fill = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

## Evidence Quality {#sec-quality}

Risk of bias ratings tell us how much to trust the results from each paradigm.
Systematic differences in quality -- for example, if one paradigm consistently lacks blinding or has high attrition -- would mean we should weight its pooled estimates differently.

```{r}
#| label: quality-comparison
#| eval: !expr nrow(quality) > 0
treatment_quality <- quality |>
  dplyr::filter(
    search_topic %in% c("treatment_cbt_psych", "treatment_biomedical"),
    domain_name == "overall"
  )

if (nrow(treatment_quality) > 0) {
  treatment_quality <- treatment_quality |>
    dplyr::mutate(paradigm = dplyr::if_else(
      search_topic == "treatment_cbt_psych",
      "Psychological / Behavioural",
      "Biomedical"
    ))

  rob_summary <- treatment_quality |>
    dplyr::count(Paradigm = paradigm, `Overall RoB` = judgement) |>
    dplyr::filter(n > 0)
  names(rob_summary) <- c("Paradigm", "Overall RoB", "Studies")
  display_table(rob_summary)
}
```

## Summary Table {#sec-summary}

The table below provides a quick reference for comparing the volume and maturity of evidence in each paradigm.
Pay attention to the number of RCTs relative to total studies -- a paradigm with many studies but few RCTs has a weaker evidence base than one with a higher proportion of randomised designs.

```{r}
#| label: summary-table
if (nrow(all_treatment) > 0) {
  design_col <- if ("llm_study_type" %in% names(all_treatment)) "llm_study_type" else "study_type"
  summary_df <- all_treatment |>
    dplyr::mutate(is_rct = grepl("RCT|random", .data[[design_col]], ignore.case = TRUE)) |>
    dplyr::group_by(paradigm) |>
    dplyr::summarise(
      Studies = dplyr::n(),
      RCTs = sum(is_rct, na.rm = TRUE),
      `Year range` = paste(range(year, na.rm = TRUE), collapse = "\u2013"),
      .groups = "drop"
    ) |>
    dplyr::rename(Paradigm = paradigm)
  display_table(summary_df)
}
```

::: {.callout-note}
## Data Status

```{r}
#| label: data-status
if (!has_cbt && !has_biomed) {
  cli::cat_line("No extracted treatment data yet. Run the search and extraction pipeline to populate this comparison.")
} else {
  n_cbt <- if (has_cbt) nrow(cbt) else 0L
  n_bio <- if (has_biomed) nrow(biomed) else 0L
  cli::cat_line(glue::glue("{n_cbt} psychological and {n_bio} biomedical effect sizes extracted."))
}
```
:::
