---
title: "Treatment: Biomedical Interventions"
aliases:
  - /reviews/biomedical/
  - /reviews/pharmacological/
---

```{r}
#| label: setup
#| include: false
source(here::here("R/utils.R"))
source(here::here("R/meta-analysis.R"))
source(here::here("R/plots.R"))

studies <- read_study_database()
effects <- read_extracted_data("treatment_biomedical")
quality <- read_quality_ratings()
has_data <- nrow(effects) > 0
```

## Overview

This review covers pharmacological and device-based interventions for Long COVID and ME/CFS, including low-dose naltrexone (LDN), antihistamines, antivirals, immunomodulators, anticoagulants, apheresis, and nutritional supplements.

```{r}
#| label: included-studies
#| results: asis
topic_studies <- studies |>
  dplyr::filter(
    !is.na(search_topic), search_topic == "treatment_biomedical",
    !is.na(llm_relevant), llm_relevant
  )
if (nrow(topic_studies) > 0) {
  n_proto <- if ("is_protocol" %in% names(topic_studies)) sum(topic_studies$is_protocol, na.rm = TRUE) else 0L
  n_results <- nrow(topic_studies) - n_proto
  n_rct <- sum(grepl("RCT|random", topic_studies$llm_study_type %||% "", ignore.case = TRUE) &
    !topic_studies$is_protocol, na.rm = TRUE)
  cli::cat_line(glue::glue("**{n_results} studies with results** ({n_rct} RCTs), plus {n_proto} registered protocols."))
}
```

## Effect Sizes {#sec-effects}

```{r}
#| label: forest-data
#| include: false
forest_fit <- NULL
forest_h <- 6
if (has_data && all(c("mean_intervention", "sd_intervention", "n_intervention",
                      "mean_comparator", "sd_comparator", "n_comparator") %in% names(effects))) {
  complete <- effects |>
    dplyr::filter(
      !is.na(mean_intervention), !is.na(mean_comparator),
      !is.na(sd_intervention), !is.na(sd_comparator),
      !is.na(n_intervention), !is.na(n_comparator),
      n_intervention > 0, n_comparator > 0,
      sd_intervention > 0, sd_comparator > 0
    ) |>
    dplyr::rename(
      mean_treatment = mean_intervention,
      sd_treatment   = sd_intervention,
      n_treatment    = n_intervention,
      mean_control   = mean_comparator,
      sd_control     = sd_comparator,
      n_control      = n_comparator
    ) |>
    dplyr::mutate(study_label = study_id)
  if (nrow(complete) >= 2) {
    es <- compute_effect_sizes(complete, outcome_type = "continuous")
    forest_fit <- run_meta(es)
    forest_h <- forest_height(forest_fit$k)
  }
}
```

```{r}
#| label: forest-plot
#| eval: !expr '!is.null(forest_fit)'
#| fig-height: !expr forest_h
#| fig-cap: "Effect sizes for biomedical interventions (positive = benefit)"
forest_gg(forest_fit, title = "Biomedical Interventions")
```

## Interventions {#sec-interventions}

```{r}
#| label: intervention-table
#| eval: !expr has_data
if (has_data && "intervention" %in% names(effects)) {
  int_counts <- effects |>
    dplyr::filter(!is.na(intervention)) |>
    dplyr::count(intervention) |>
    dplyr::arrange(dplyr::desc(n))
  if (nrow(int_counts) > 0) {
    names(int_counts) <- c("Intervention", "Effect Sizes")
    display_table(int_counts)
  }
}
```

## Dosing & Protocol {#sec-dose}

```{r}
#| label: dose-table
#| eval: !expr has_data
if (has_data && "dose" %in% names(effects)) {
  dose_df <- effects |>
    dplyr::filter(!is.na(dose), !is.na(intervention)) |>
    dplyr::select(study_id, intervention, dose) |>
    dplyr::distinct()
  if (nrow(dose_df) > 0) {
    names(dose_df) <- c("Study", "Intervention", "Dose / Protocol")
    display_table(dose_df)
  }
}
```

## PEM Assessment {#sec-pem}

```{r}
#| label: pem-status
#| results: asis
#| eval: !expr has_data
if (has_data && "pem_assessed" %in% names(effects)) {
  n_pem <- sum(effects$pem_assessed, na.rm = TRUE)
  pct <- round(n_pem / nrow(effects) * 100)
  cli::cat_line(glue::glue("PEM was assessed in **{n_pem}/{nrow(effects)}** ({pct}%) reported outcomes."))
}
```

## Risk of Bias {#sec-rob}

```{r}
#| label: rob-summary
#| eval: !expr nrow(quality) > 0
biomed_quality <- quality |>
  dplyr::filter(!is.na(search_topic), search_topic == "treatment_biomedical", domain_name == "overall")
if (nrow(biomed_quality) > 0) {
  rob_counts <- biomed_quality |>
    dplyr::count(judgement)
  names(rob_counts) <- c("Overall RoB", "Studies")
  display_table(rob_counts)
}
```

## Adverse Events {#sec-safety}

```{r}
#| label: adverse-events
#| results: asis
#| eval: !expr has_data
if (has_data && "adverse_events_reported" %in% names(effects)) {
  n_ae <- sum(effects$adverse_events_reported, na.rm = TRUE)
  pct <- if (nrow(effects) > 0) round(n_ae / nrow(effects) * 100) else 0
  cli::cat_line(glue::glue("Adverse events reported in **{n_ae}/{nrow(effects)}** ({pct}%) outcomes."))
  if (n_ae > 0 && "adverse_events_detail" %in% names(effects)) {
    details <- effects$adverse_events_detail[!is.na(effects$adverse_events_detail) & effects$adverse_events_detail != ""]
    if (length(details) > 0) {
      cli::cat_line("\n**Reported adverse events:**\n")
      purrr::walk(unique(details), \(d) cli::cat_bullet(d))
    }
  }
}
```

::: {.callout-note}
## Data Status

```{r}
#| label: data-status
#| results: asis
if (!has_data) {
  cli::cat_line("No extracted data yet. Run the extraction pipeline to populate this review.")
} else {
  n_studies <- dplyr::n_distinct(effects$study_id)
  cli::cat_line(glue::glue("{nrow(effects)} effect sizes from {n_studies} studies."))
}
```
:::
