---
title: "Treatment: Biomedical Interventions"
aliases:
  - /reviews/biomedical/
  - /reviews/pharmacological/
---

```{r}
#| label: setup
#| include: false
source(here::here("R/utils.R"))
source(here::here("R/meta-analysis.R"))
source(here::here("R/plots.R"))

studies <- read_study_database()
effects <- read_extracted_data("treatment_biomedical")
quality <- read_quality_ratings()
has_data <- nrow(effects) > 0
```

## Overview

Biomedical interventions for Long COVID and ME/CFS span a wide range — from repurposed pharmaceuticals like low-dose naltrexone and antivirals, to nutritional supplements like CoQ10 and creatine, to procedural treatments like apheresis and transcranial stimulation.

The evidence is growing fast but remains fragmented.
Most studies are small, many lack placebo controls, and the interventions target different mechanisms entirely.
Pooling across such different treatments is statistically possible but clinically questionable, so we present both the overall picture and domain-specific breakdowns below.

```{r}
#| label: included-studies
#| results: asis
topic_studies <- studies |>
  dplyr::filter(
    !is.na(search_topic), search_topic == "treatment_biomedical",
    !is.na(llm_relevant), llm_relevant
  )
if (nrow(topic_studies) > 0) {
  n_proto <- if ("is_protocol" %in% names(topic_studies)) sum(topic_studies$is_protocol, na.rm = TRUE) else 0L
  n_results <- nrow(topic_studies) - n_proto
  n_rct <- sum(grepl("RCT|random", topic_studies$llm_study_type %||% "", ignore.case = TRUE) &
    !topic_studies$is_protocol, na.rm = TRUE)
  cli::cat_line(glue::glue("**{n_results} studies with results** ({n_rct} RCTs), plus {n_proto} registered protocols."))
}
```

## Effects by Intervention {#sec-effects-intervention}

Pooling LDN with nirmatrelvir with creatine into a single number would be meaningless.
The plot below shows a separate pooled estimate for each intervention category — answering the question "what does the evidence say about _this specific_ treatment?"

```{r}
#| label: forest-data
#| include: false
forest_df <- if (has_data) prepare_forest_data(effects) else data.frame()
has_forest <- nrow(forest_df) >= 2
intervention_forest_h <- 6
domain_forest_h <- 6
if (has_forest && "intervention_category" %in% names(forest_df)) {
  n_int <- length(unique(forest_df$intervention_category[!is.na(forest_df$intervention_category)]))
  intervention_forest_h <- bar_height(n_int)
}
if (has_forest && "outcome_domain" %in% names(forest_df)) {
  n_dom <- length(unique(forest_df$outcome_domain[!is.na(forest_df$outcome_domain)]))
  domain_forest_h <- bar_height(n_dom)
}
```

```{r}
#| label: intervention-forest
#| eval: !expr 'has_forest && "intervention_category" %in% names(forest_df)'
#| fig-height: !expr intervention_forest_h
#| fig-alt: "Forest plot showing pooled effect sizes for each biomedical intervention category, with diamond size proportional to the number of contributing studies"
#| fig-cap: "Pooled effect sizes by intervention. Each row is a separate random-effects meta-analysis. Larger diamonds indicate more contributing study–outcome pairs."
subgroup_forest(
  forest_df, group_col = "intervention_category",
  title = "Pooled effects by intervention"
)
```

## Effects by Outcome Domain {#sec-effects-domain}

Same evidence, different cut.
This groups by what was measured — fatigue, cognition, pain — regardless of which drug or supplement was used.
Useful for asking: "across all biomedical approaches, which symptoms are most responsive to treatment?"

```{r}
#| label: domain-forest
#| eval: !expr 'has_forest && "outcome_domain" %in% names(forest_df)'
#| fig-height: !expr domain_forest_h
#| fig-alt: "Forest plot showing pooled effect sizes for each outcome domain across biomedical interventions"
#| fig-cap: "Pooled effect sizes by outcome domain. Each row pools all biomedical interventions that measured that domain."
subgroup_forest(
  forest_df, group_col = "outcome_domain",
  title = "Pooled effects by outcome domain"
)
```

## Interventions {#sec-interventions}

The table below groups specific drugs, supplements, and procedures into broader categories.
Low-dose naltrexone and nirmatrelvir/ritonavir have the most reported outcomes, though many of those come from observational rather than randomised studies.

```{r}
#| label: intervention-table
#| eval: !expr has_data
if (has_data && "intervention_category" %in% names(effects)) {
  int_counts <- effects |>
    dplyr::filter(!is.na(intervention_category)) |>
    dplyr::count(intervention_category, sort = TRUE)
  if (nrow(int_counts) > 0) {
    names(int_counts) <- c("Intervention", "Effect Sizes")
    display_table(int_counts)
  }
}
```

## Outcome Measures {#sec-outcomes}

As with psychological interventions, outcome measurement varies widely.
The table below maps each instrument to its domain — making it easier to spot that, for example, three different fatigue scales are measuring the same underlying construct.

```{r}
#| label: outcome-table
#| eval: !expr has_data
if (has_data && "outcome_measure" %in% names(effects)) {
  out_counts <- effects |>
    dplyr::filter(!is.na(outcome_measure)) |>
    dplyr::mutate(
      domain = dplyr::coalesce(outcome_domain, "other")
    ) |>
    dplyr::count(domain, outcome_measure, sort = TRUE)
  if (nrow(out_counts) > 0) {
    names(out_counts) <- c("Domain", "Outcome Measure", "N")
    display_table(out_counts)
  }
}
```

## Outcomes by Domain {#sec-domains}

Grouping by outcome domain cuts through the instrument heterogeneity.
The bar chart shows where the evidence concentrates; the summary table pools effect sizes within each domain.

High I^2^ values indicate substantial heterogeneity within a domain — expected, given how different the interventions are.

```{r}
#| label: domain-bar-data
#| include: false
domain_df <- NULL
domain_h <- 4
if (has_data && "outcome_domain" %in% names(effects)) {
  domain_df <- effects |>
    dplyr::filter(!is.na(outcome_domain)) |>
    dplyr::count(outcome_domain, sort = TRUE) |>
    dplyr::mutate(outcome_domain = factor(outcome_domain, levels = rev(outcome_domain)))
  domain_h <- bar_height(nrow(domain_df))
}
```

```{r}
#| label: domain-bar
#| eval: !expr '!is.null(domain_df) && nrow(domain_df) > 0'
#| fig-height: !expr domain_h
#| fig-alt: "Horizontal bar chart showing the number of extracted effect sizes per outcome domain for biomedical interventions"
#| fig-cap: "Distribution of effect sizes across outcome domains. Fatigue dominates, reflecting both clinical priority and measurement availability."
ggplot(domain_df, aes(x = n, y = outcome_domain)) +
  geom_col(fill = cols_review["clinical"], width = 0.7) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Effect sizes", y = NULL) +
  theme_review
```

```{r}
#| label: domain-forest-data
#| include: false
domain_fits <- list()
domain_forest_h <- 6
if (has_data && all(c("mean_intervention", "sd_intervention", "n_intervention",
                      "mean_comparator", "sd_comparator", "n_comparator",
                      "outcome_domain") %in% names(effects))) {
  for (dom in unique(effects$outcome_domain[!is.na(effects$outcome_domain)])) {
    dom_data <- effects |>
      dplyr::filter(
        outcome_domain == dom,
        !is.na(mean_intervention), !is.na(mean_comparator),
        !is.na(sd_intervention), !is.na(sd_comparator),
        !is.na(n_intervention), !is.na(n_comparator),
        n_intervention > 0, n_comparator > 0,
        sd_intervention > 0, sd_comparator > 0
      ) |>
      dplyr::rename(
        mean_treatment = mean_intervention, sd_treatment = sd_intervention,
        n_treatment = n_intervention, mean_control = mean_comparator,
        sd_control = sd_comparator, n_control = n_comparator
      ) |>
      dplyr::mutate(study_label = paste0(study_id, ": ", outcome_measure))
    if (nrow(dom_data) >= 2) {
      es <- compute_effect_sizes(dom_data, outcome_type = "continuous")
      domain_fits[[dom]] <- run_meta(es)
    }
  }
  if (length(domain_fits) > 0) {
    total_k <- sum(vapply(domain_fits, \(f) f$k, integer(1)))
    domain_forest_h <- forest_height(total_k + length(domain_fits))
  }
}
```

```{r}
#| label: domain-summary-table
#| eval: !expr 'length(domain_fits) > 0'
domain_summary <- purrr::imap_dfr(domain_fits, \(fit, dom) {
  data.frame(
    Domain = dom,
    k = fit$k,
    SMD = round(coef(fit), 2),
    CI = paste0("[", round(fit$ci.lb, 2), ", ", round(fit$ci.ub, 2), "]"),
    p = formatC(fit$pval, format = "g", digits = 3),
    I2 = paste0(round(fit$I2, 1), "%")
  )
}) |> dplyr::arrange(SMD)
display_table(domain_summary)
```

## Dosing & Protocol {#sec-dose}

Dosing information matters for biomedical interventions in a way it does not for CBT.
The table below lists each study's reported dose or protocol, grouped by intervention category.

```{r}
#| label: dose-table
#| eval: !expr has_data
if (has_data && "dose" %in% names(effects)) {
  dose_df <- effects |>
    dplyr::filter(!is.na(dose), !is.na(intervention_category)) |>
    dplyr::select(study_id, intervention_category, dose) |>
    dplyr::distinct()
  if (nrow(dose_df) > 0) {
    names(dose_df) <- c("Study", "Intervention", "Dose / Protocol")
    display_table(dose_df)
  }
}
```

## PEM Assessment {#sec-pem}

Post-exertional malaise (PEM) is the hallmark symptom of ME/CFS and a defining feature of Long COVID for many patients.
Whether a biomedical trial assessed PEM matters less for safety than in exercise trials, but it is still relevant — some pharmacological interventions aim specifically to reduce PEM, and knowing whether they measured it tells us whether they could detect their own target outcome.

```{r}
#| label: pem-status
#| results: asis
#| eval: !expr has_data
if (has_data && "pem_assessed" %in% names(effects)) {
  n_pem <- sum(effects$pem_assessed, na.rm = TRUE)
  pct <- round(n_pem / nrow(effects) * 100)
  cli::cat_line(glue::glue("PEM was assessed in **{n_pem}/{nrow(effects)}** ({pct}%) reported outcomes."))
}
```

## Risk of Bias {#sec-rob}

Risk of bias ratings follow the Cochrane ROB2 tool for RCTs and ROBINS-I for observational studies.

```{r}
#| label: rob-summary
#| eval: !expr nrow(quality) > 0
biomed_quality <- quality |>
  dplyr::filter(!is.na(search_topic), search_topic == "treatment_biomedical", domain_name == "overall")
if (nrow(biomed_quality) > 0) {
  rob_counts <- biomed_quality |>
    dplyr::count(judgement)
  names(rob_counts) <- c("Overall RoB", "Studies")
  display_table(rob_counts)
}
```

## Adverse Events {#sec-safety}

Adverse event reporting is inconsistent.
Some studies report detailed safety data; many say nothing at all.
Absence of adverse event reporting does not mean absence of adverse events.

```{r}
#| label: adverse-events
#| results: asis
#| eval: !expr has_data
if (has_data && "adverse_events_reported" %in% names(effects)) {
  n_ae <- sum(effects$adverse_events_reported, na.rm = TRUE)
  pct <- if (nrow(effects) > 0) round(n_ae / nrow(effects) * 100) else 0
  cli::cat_line(glue::glue("Adverse events reported in **{n_ae}/{nrow(effects)}** ({pct}%) outcomes."))
  if (n_ae > 0 && "adverse_events_detail" %in% names(effects)) {
    details <- effects$adverse_events_detail[!is.na(effects$adverse_events_detail) & effects$adverse_events_detail != ""]
    if (length(details) > 0) {
      cli::cat_line("\n**Reported adverse events:**\n")
      purrr::walk(unique(details), \(d) cli::cat_bullet(d))
    }
  }
}
```

::: {.callout-note}
## Data Status

```{r}
#| label: data-status
#| results: asis
if (!has_data) {
  cli::cat_line("No extracted data yet. Run the extraction pipeline to populate this review.")
} else {
  n_studies <- dplyr::n_distinct(effects$study_id)
  cli::cat_line(glue::glue("{nrow(effects)} effect sizes from {n_studies} studies."))
}
```

This is a living review.
Data is extracted automatically from published studies and updated as new evidence becomes available.
:::
