---
title: "Treatment: CBT & Psychological Interventions"
aliases:
  - /reviews/cbt/
  - /reviews/psychological/
---

```{r}
#| label: setup
#| include: false
source(here::here("R/utils.R"))
source(here::here("R/meta-analysis.R"))
source(here::here("R/plots.R"))

studies <- read_study_database()
effects <- read_extracted_data("treatment_cbt_psych")
quality <- read_quality_ratings()
has_data <- nrow(effects) > 0
```

## Overview

Psychological and behavioural interventions are among the most studied — and most debated — approaches for Long COVID and ME/CFS.
This review covers CBT, graded exercise therapy (GET), acceptance and commitment therapy (ACT), the Lightning Process, mindfulness, pacing programmes, and self-management interventions.

The evidence base is large, but the picture is complicated.
Many trials use different outcome measures, different diagnostic criteria, and different definitions of what counts as "improvement."
The data below reflects that heterogeneity honestly.

```{r}
#| label: included-studies
#| results: asis
topic_studies <- studies |>
  dplyr::filter(
    !is.na(search_topic), search_topic == "treatment_cbt_psych",
    !is.na(llm_relevant), llm_relevant
  )
if (nrow(topic_studies) > 0) {
  n_proto <- if ("is_protocol" %in% names(topic_studies)) sum(topic_studies$is_protocol, na.rm = TRUE) else 0L
  n_results <- nrow(topic_studies) - n_proto
  n_rct <- sum(grepl("RCT|random", topic_studies$llm_study_type %||% "", ignore.case = TRUE) &
    !topic_studies$is_protocol, na.rm = TRUE)
  cli::cat_line(glue::glue("**{n_results} studies with results** ({n_rct} RCTs), plus {n_proto} registered protocols."))
}
```

## Effects by Intervention {#sec-effects-intervention}

Pooling every psychological intervention into a single number hides more than it reveals.
The plot below breaks the evidence apart by intervention category — each diamond is the pooled effect for one type of intervention, sized by the number of study–outcome pairs contributing.
This tells you which approaches have been studied most and where the effect estimates point.

```{r}
#| label: forest-data
#| include: false
forest_df <- if (has_data) prepare_forest_data(effects) else data.frame()
has_forest <- nrow(forest_df) >= 2
intervention_forest_h <- 6
domain_forest_h <- 6
if (has_forest && "intervention_category" %in% names(forest_df)) {
  n_int <- length(unique(forest_df$intervention_category[!is.na(forest_df$intervention_category)]))
  intervention_forest_h <- bar_height(n_int)
}
if (has_forest && "outcome_domain" %in% names(forest_df)) {
  n_dom <- length(unique(forest_df$outcome_domain[!is.na(forest_df$outcome_domain)]))
  domain_forest_h <- bar_height(n_dom)
}
```

```{r}
#| label: intervention-forest
#| eval: !expr 'has_forest && "intervention_category" %in% names(forest_df)'
#| fig-height: !expr intervention_forest_h
#| fig-alt: "Forest plot showing pooled effect sizes for each intervention category, with diamond size proportional to the number of contributing studies"
#| fig-cap: "Pooled effect sizes by intervention type. Each row is a separate random-effects meta-analysis. Larger diamonds indicate more contributing study–outcome pairs. Negative values favour the intervention over the comparator."
subgroup_forest(
  forest_df, group_col = "intervention_category",
  title = "Pooled effects by intervention type"
)
```

## Effects by Outcome Domain {#sec-effects-domain}

The same data, sliced differently.
Here the pooling is by what was measured rather than what was done.
This answers a different question: across all psychological interventions, which outcomes show the most consistent effects?

```{r}
#| label: domain-forest
#| eval: !expr 'has_forest && "outcome_domain" %in% names(forest_df)'
#| fig-height: !expr domain_forest_h
#| fig-alt: "Forest plot showing pooled effect sizes for each outcome domain, with diamond size proportional to the number of contributing studies"
#| fig-cap: "Pooled effect sizes by outcome domain. Each row pools all interventions that measured that domain. Negative values indicate improvement."
subgroup_forest(
  forest_df, group_col = "outcome_domain",
  title = "Pooled effects by outcome domain"
)
```

## Interventions {#sec-interventions}

The table below groups individual intervention names into broader categories.
A single study may contribute multiple effect sizes — one per outcome — so the counts reflect reported outcomes, not unique trials.

CBT and exercise-based approaches dominate the literature, but the category labels hide real diversity.
"CBT" includes everything from face-to-face individual therapy to internet-delivered group programmes.
"Exercise" spans supervised gym sessions to home-based walking programmes.

```{r}
#| label: intervention-table
#| eval: !expr has_data
if (has_data && "intervention_category" %in% names(effects)) {
  int_counts <- effects |>
    dplyr::filter(!is.na(intervention_category)) |>
    dplyr::count(intervention_category, sort = TRUE)
  if (nrow(int_counts) > 0) {
    names(int_counts) <- c("Intervention", "Effect Sizes")
    display_table(int_counts)
  }
}
```

## Outcome Measures {#sec-outcomes}

Outcome measurement is one of the biggest sources of heterogeneity in this literature.
The table below shows which instruments are used, grouped by the domain they measure.
The same construct — fatigue, say — gets measured with the Chalder Fatigue Scale in one trial, the MFI-20 in another, and a simple VAS in a third.
This makes direct comparison across studies harder than it should be.

```{r}
#| label: outcome-table
#| eval: !expr has_data
if (has_data && "outcome_measure" %in% names(effects)) {
  out_counts <- effects |>
    dplyr::filter(!is.na(outcome_measure)) |>
    dplyr::mutate(
      domain = dplyr::coalesce(outcome_domain, "other")
    ) |>
    dplyr::count(domain, outcome_measure, sort = TRUE)
  if (nrow(out_counts) > 0) {
    names(out_counts) <- c("Domain", "Outcome Measure", "N")
    display_table(out_counts)
  }
}
```

## Outcomes by Domain {#sec-domains}

Rather than comparing individual instruments, we can group outcomes into broad domains — fatigue, physical function, quality of life, and so on — to see where the evidence concentrates.

The bar chart shows how many effect sizes fall into each domain.
The summary table below it pools effect sizes within each domain that has enough studies for a meta-analysis (at least two with complete data).

```{r}
#| label: domain-bar-data
#| include: false
domain_df <- NULL
domain_h <- 4
if (has_data && "outcome_domain" %in% names(effects)) {
  domain_df <- effects |>
    dplyr::filter(!is.na(outcome_domain)) |>
    dplyr::count(outcome_domain, sort = TRUE) |>
    dplyr::mutate(outcome_domain = factor(outcome_domain, levels = rev(outcome_domain)))
  domain_h <- bar_height(nrow(domain_df))
}
```

```{r}
#| label: domain-bar
#| eval: !expr '!is.null(domain_df) && nrow(domain_df) > 0'
#| fig-height: !expr domain_h
#| fig-alt: "Horizontal bar chart showing the number of extracted effect sizes per outcome domain, with fatigue and physical function having the most"
#| fig-cap: "Distribution of effect sizes across outcome domains. Each bar represents how many study–outcome pairs reported on that domain."
ggplot(domain_df, aes(x = n, y = outcome_domain)) +
  geom_col(fill = cols_review["clinical"], width = 0.7) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Effect sizes", y = NULL) +
  theme_review
```

```{r}
#| label: domain-forest-data
#| include: false
domain_fits <- list()
domain_forest_h <- 6
if (has_data && all(c("mean_intervention", "sd_intervention", "n_intervention",
                      "mean_comparator", "sd_comparator", "n_comparator",
                      "outcome_domain") %in% names(effects))) {
  for (dom in unique(effects$outcome_domain[!is.na(effects$outcome_domain)])) {
    dom_data <- effects |>
      dplyr::filter(
        outcome_domain == dom,
        !is.na(mean_intervention), !is.na(mean_comparator),
        !is.na(sd_intervention), !is.na(sd_comparator),
        !is.na(n_intervention), !is.na(n_comparator),
        n_intervention > 0, n_comparator > 0,
        sd_intervention > 0, sd_comparator > 0
      ) |>
      dplyr::rename(
        mean_treatment = mean_intervention, sd_treatment = sd_intervention,
        n_treatment = n_intervention, mean_control = mean_comparator,
        sd_control = sd_comparator, n_control = n_comparator
      ) |>
      dplyr::mutate(study_label = paste0(study_id, ": ", outcome_measure))
    if (nrow(dom_data) >= 2) {
      es <- compute_effect_sizes(dom_data, outcome_type = "continuous")
      domain_fits[[dom]] <- run_meta(es)
    }
  }
  if (length(domain_fits) > 0) {
    total_k <- sum(vapply(domain_fits, \(f) f$k, integer(1)))
    domain_forest_h <- forest_height(total_k + length(domain_fits))
  }
}
```

```{r}
#| label: domain-summary-table
#| eval: !expr 'length(domain_fits) > 0'
domain_summary <- purrr::imap_dfr(domain_fits, \(fit, dom) {
  data.frame(
    Domain = dom,
    k = fit$k,
    SMD = round(coef(fit), 2),
    CI = paste0("[", round(fit$ci.lb, 2), ", ", round(fit$ci.ub, 2), "]"),
    p = formatC(fit$pval, format = "g", digits = 3),
    I2 = paste0(round(fit$I2, 1), "%")
  )
}) |> dplyr::arrange(SMD)
display_table(domain_summary)
```

## PEM Assessment {#sec-pem}

Post-exertional malaise (PEM) is the hallmark symptom of ME/CFS and a core feature of Long COVID.
Whether a study assessed PEM matters because interventions that increase activity — like GET — can trigger PEM in susceptible patients.
A trial that does not measure PEM cannot detect this harm.

```{r}
#| label: pem-status
#| results: asis
#| eval: !expr has_data
if (has_data && "pem_assessed" %in% names(effects)) {
  n_pem <- sum(effects$pem_assessed, na.rm = TRUE)
  n_total <- nrow(effects)
  pct <- round(n_pem / n_total * 100)
  cli::cat_line(glue::glue("PEM was assessed in **{n_pem}/{n_total}** ({pct}%) reported outcomes."))
}
```

## Risk of Bias {#sec-rob}

Risk of bias ratings follow the Cochrane ROB2 tool for RCTs and ROBINS-I for observational studies.
The table below shows overall judgements where quality assessment has been completed.

```{r}
#| label: rob-summary
#| eval: !expr nrow(quality) > 0
psych_quality <- quality |>
  dplyr::filter(!is.na(search_topic), search_topic == "treatment_cbt_psych",
                domain_name == "overall")
if (nrow(psych_quality) > 0) {
  rob_counts <- psych_quality |>
    dplyr::count(judgement)
  names(rob_counts) <- c("Overall RoB", "Studies")
  display_table(rob_counts)
}
```

## Adverse Events {#sec-safety}

Adverse event reporting is inconsistent across trials.
Many studies do not mention adverse events at all, which does not mean none occurred — it means we cannot tell.

```{r}
#| label: adverse-events
#| results: asis
#| eval: !expr has_data
if (has_data && "adverse_events_reported" %in% names(effects)) {
  n_ae <- sum(effects$adverse_events_reported, na.rm = TRUE)
  cli::cat_line(glue::glue("Adverse events reported in **{n_ae}/{nrow(effects)}** outcomes."))
}
```

::: {.callout-note}
## Data Status

```{r}
#| label: data-status
#| results: asis
if (!has_data) {
  cli::cat_line("No extracted data yet. Run the extraction pipeline to populate this review.")
} else {
  n_studies <- dplyr::n_distinct(effects$study_id)
  cli::cat_line(glue::glue("{nrow(effects)} effect sizes from {n_studies} studies."))
}
```

This is a living review.
Data is extracted automatically from published studies and updated as new evidence becomes available.
:::
