---
title: "Treatment Effectiveness"
---

```{r}
#| label: setup
#| include: false
source("../R/utils.R")
source("../R/meta-analysis.R")
source("../R/plots.R")

studies <- read_study_database()
effects <- read_extracted_data("treatment_effectiveness")
quality <- read_quality_ratings()
has_data <- nrow(effects) > 0
```

## Overview

Pooling treatment effects across heterogeneous interventions -- pacing alongside pharmacology, CBT alongside supplements -- gives a bird's-eye view of the field, but it is deliberately broad.
The pooled estimate answers a general question: across everything that has been tried, is there a detectable signal of benefit?
That number matters less than the variation around it: high heterogeneity tells us that "what works" depends heavily on which intervention, which population, and which outcome you look at.

This review examines the comparative effectiveness of interventions for Long COVID and ME/CFS, including:

- **Pacing / energy management** vs usual care or other active interventions
- **Graded exercise therapy (GET)** vs usual care or pacing
- **Cognitive behavioural therapy (CBT)** vs usual care
- **Pharmacological interventions** (LDN, antihistamines, beta-blockers, etc.)
- **Supplements** (CoQ10, NAD+, etc.)

Clinical and patient-reported outcome studies are analysed separately because they answer different questions and carry different biases.

```{r}
#| label: included-studies
#| eval: !expr has_data
treatment_studies <- studies[studies$search_topic == "treatment_effectiveness" & studies$included, ]
cat(sprintf("**%d studies included** in treatment effectiveness review.\n", nrow(treatment_studies)))
```

## Clinical Studies {#sec-clinical}

Clinical outcomes -- measured by clinicians, laboratory tests, or objective performance assessments -- tend to be less susceptible to expectation bias than self-report measures.
When a treatment shows an effect on clinical outcomes, it provides stronger evidence that something physiological is changing.
The forest plot below displays individual study effects alongside the pooled estimate, so you can see both the overall signal and how much studies disagree.

```{r}
#| label: forest-clinical
#| eval: !expr has_data
#| fig-alt: "Forest plot showing Hedges' g effect sizes for individual clinical studies, with a pooled diamond estimate at the bottom."
#| fig-cap: "Forest plot of treatment effects from clinical studies (Hedges' g)"
clinical <- effects[effects$reporting_type == "clinical", ]
if (nrow(clinical) >= 2) {
  es <- compute_effect_sizes(clinical, outcome_type = "continuous")
  fit <- run_meta(es)
  forest_gg(fit, title = "Treatment Effects \u2014 Clinical Studies")
}
```

```{r}
#| label: heterogeneity-clinical
#| eval: !expr has_data
if (nrow(clinical) >= 2) {
  het <- heterogeneity_summary(fit)
  cat(sprintf(
    "**Heterogeneity**: Q(%d) = %.1f, p = %.3f; I\u00b2 = %.1f%%; \u03c4\u00b2 = %.3f\n",
    het$Q_df, het$Q, het$Q_p, het$I2, het$tau2
  ))
}
```

### Subgroup: By PEM Status

Post-exertional malaise (PEM) is a cardinal feature of ME/CFS and common in Long COVID.
Trials that confirm PEM in their participants may be studying a biologically distinct population.
If PEM-confirmed cohorts respond differently to treatment -- especially to exercise-based interventions -- that tells us PEM status should be a standard stratification variable in future trials.

```{r}
#| label: subgroup-pem-clinical
#| eval: !expr has_data
#| fig-alt: "Forest plot of treatment effects stratified by whether the study population had confirmed post-exertional malaise."
#| fig-cap: "Treatment effects stratified by PEM status"
if (nrow(clinical) >= 2 && any(!is.na(clinical$pem_status))) {
  fit_sub <- run_subgroup_meta(es, moderator = "pem_status")
  cat(sprintf("Test of moderators: F = %.2f, p = %.3f\n", fit_sub$QM, fit_sub$QMp))
}
```

## Patient-Reported Studies {#sec-patient-reported}

Patient-reported outcomes capture what matters most to the people living with these conditions: fatigue severity, functional capacity, quality of life.
They are also more vulnerable to placebo effects and unblinding, which is why we analyse them separately from clinical measures.
A treatment that improves patient-reported outcomes without moving clinical markers may reflect genuine subjective benefit, a placebo response, or a real effect that our clinical tools are too blunt to detect.

```{r}
#| label: forest-patient
#| eval: !expr has_data
#| fig-alt: "Forest plot showing Hedges' g effect sizes for individual patient-reported outcome studies, with a pooled diamond estimate."
#| fig-cap: "Forest plot of treatment effects from patient-reported studies"
patient <- effects[effects$reporting_type == "patient_reported", ]
if (nrow(patient) >= 2) {
  es_pat <- compute_effect_sizes(patient, outcome_type = "continuous")
  fit_pat <- run_meta(es_pat)
  forest_gg(fit_pat, title = "Treatment Effects \u2014 Patient-Reported Studies")
}
```

## Lay Summary {#sec-lay}

```{r}
#| label: lay-summary
#| eval: !expr has_data
#| fig-alt: "Simplified visualisation summarising treatment findings for a general audience."
#| fig-cap: "What patients and clinicians report about treatments"
cat("*This section provides accessible visualisations for non-specialist audiences.*\n\n")
```

## Study Quality {#sec-quality}

Risk of bias and GRADE ratings determine how much weight the pooled estimates deserve.
A meta-analysis built on studies with high risk of bias will produce a precise but potentially misleading number.
The traffic light plot below shows domain-level bias judgements for each study, and the GRADE heatmap summarises overall evidence certainty for each comparison.

```{r}
#| label: rob-plot
#| eval: !expr nrow(quality) > 0
#| fig-alt: "Traffic light plot showing risk of bias judgements across multiple domains for each included treatment study."
#| fig-cap: "Risk of bias assessment for included treatment studies"
treatment_quality <- quality[quality$comparison != "", ]
if (nrow(treatment_quality) > 0) {
  rob_traffic_light(treatment_quality)
}
```

```{r}
#| label: grade-plot
#| eval: !expr nrow(quality) > 0
#| fig-alt: "Heatmap showing GRADE evidence certainty ratings for each treatment comparison, coloured from very low to high."
#| fig-cap: "GRADE evidence certainty ratings"
if (nrow(treatment_quality) > 0 && any(!is.na(treatment_quality$grade))) {
  grade_heatmap(treatment_quality)
}
```

## Publication Bias {#sec-pub-bias}

Funnel plot asymmetry can suggest that small studies with negative results are missing from the evidence base.
If the plot shows a gap in the lower-left region (small studies with small or negative effects), that is a warning sign: the pooled estimate may be inflated because unfavourable results were never published or were published in harder-to-find venues.
Asymmetry alone does not prove publication bias -- it can also reflect genuine heterogeneity or differences in study quality -- but it warrants caution.

```{r}
#| label: funnel-plot
#| eval: !expr has_data
#| fig-alt: "Funnel plot showing individual study effect sizes against their standard errors, with the pooled estimate marked as a vertical line."
#| fig-cap: "Funnel plot for treatment effectiveness studies"
if (exists("fit") && fit$k >= 5) {
  funnel_gg(fit)
}
```

## Sensitivity Analysis {#sec-sensitivity}

Leave-one-out analysis removes each study in turn and re-estimates the pooled effect.
If the overall result shifts substantially when a single study is dropped, that study is driving the conclusion and deserves closer scrutiny.
A robust finding holds up regardless of which study is excluded.

```{r}
#| label: leave-one-out
#| eval: !expr has_data
if (exists("fit") && fit$k >= 3) {
  l1o <- leave_one_out(fit)
  print(l1o)
}
```

::: {.callout-note}
## Data Status

```{r}
#| label: data-status
if (!has_data) {
  cat("No extracted treatment data yet. Run the search and extraction pipeline to populate this review.\n")
} else {
  cat(sprintf(
    "%d effect sizes from %d studies. Last updated: %s\n",
    nrow(effects),
    length(unique(effects$study_id)),
    max(studies$date_added[studies$search_topic == "treatment_effectiveness"], na.rm = TRUE)
  ))
}
```
:::

::: {.callout-tip}
## Living Review

This page is part of a living systematic review.
New studies are incorporated as they are identified through automated search updates, and the meta-analytic estimates on this page will shift as the evidence base grows.
If you are reading a cached version, check the data status note above for the last update date.
:::
