---
title: "Treatment Effectiveness"
subtitle: "Pooled evidence across all intervention types"
---

```{r}
#| label: setup
#| include: false
source("../R/utils.R")
source("../R/meta-analysis.R")
source("../R/plots.R")

studies <- read_study_database()
effects <- read_extracted_data("treatment_effectiveness")
quality <- read_quality_ratings()
has_data <- nrow(effects) > 0
```

```{r}
#| label: data-pending
#| results: asis
#| echo: false
#| eval: !expr '!has_data'
cat("::: {.callout-note}\n")
cat("## Data not yet available\n\n")
cat("The treatment effectiveness data extraction has not been completed yet. ")
cat("This review will be populated once the extraction pipeline includes the `treatment_effectiveness` domain. ")
cat("In the meantime, see the [CBT & Psychological](treatment-cbt-psych.qmd) and [Biomedical](treatment-biomedical.qmd) treatment reviews for available evidence.\n")
cat(":::\n")
```

::: {.lead}
Pooling treatment effects across heterogeneous interventions — pacing alongside pharmacology, CBT alongside supplements — gives a bird's-eye view of the field, but it is deliberately broad.
The pooled estimate answers a general question: across everything that has been tried, is there a detectable signal of benefit?
:::

```{r}
#| label: included-studies
#| results: asis
#| echo: false
#| eval: !expr has_data
treatment_studies <- studies[studies$search_topic == "treatment_effectiveness" & studies$included, ]
n <- nrow(treatment_studies)
cat(sprintf('<div class="study-count"><span class="count-number">%d</span><span class="count-label">studies included in<br>treatment review</span></div>\n', n))
```

That number matters less than the variation around it: high heterogeneity tells us that "what works" depends heavily on which intervention, which population, and which outcome you look at.

This review examines the comparative effectiveness of interventions for Long COVID and ME/CFS, including:

- **Pacing / energy management** vs usual care or other active interventions
- **Graded exercise therapy (GET)** vs usual care or pacing
- **Cognitive behavioural therapy (CBT)** vs usual care
- **Pharmacological interventions** (LDN, antihistamines, beta-blockers, etc.)
- **Supplements** (CoQ10, NAD+, etc.)

Clinical and patient-reported outcome studies are analysed separately because they answer different questions and carry different biases.

::: {.at-a-glance}

## At a Glance {#sec-lay}

These charts present the evidence in formats designed to be accessible without statistical training.

```{r}
#| label: lay-lollipop
#| eval: !expr has_data
#| fig-height: 8
#| fig-alt: "Lollipop chart showing treatment effects with plain-language labels like small, moderate, or large benefit."
#| fig-cap: "How much does each type of treatment help? Dot position shows effect size; labels translate the statistical result into plain language."
forest_df_all <- if (has_data) prepare_forest_data(effects) else data.frame()
if (nrow(forest_df_all) >= 2 && "intervention_category" %in% names(forest_df_all)) {
  lay_lollipop(forest_df_all, group_col = "intervention_category")
}
```

```{r}
#| label: lay-treatment-verdict
#| eval: !expr has_data
#| fig-height: 8
#| fig-alt: "Dot plot showing whether each treatment is likely helpful, possibly helpful, uncertain, or may not help, using traffic-light colours."
#| fig-cap: "At a glance: does the treatment help? Green dots mean the evidence points towards benefit. Grey means we cannot tell yet."
if (nrow(forest_df_all) >= 2 && "intervention_category" %in% names(forest_df_all)) {
  lay_treatment_summary(forest_df_all, group_col = "intervention_category")
}
```

```{r}
#| label: lay-grade-bar
#| eval: !expr has_data
#| fig-height: 4
#| fig-alt: "Stacked bar chart showing how many outcomes have very low, low, moderate, or high certainty evidence."
#| fig-cap: "How confident are we in these results? Most treatment evidence is rated as very uncertain or uncertain, meaning we need more and better studies."
grade <- read_grade_data()
if (nrow(grade) > 0) {
  lay_grade_bar(grade)
}
```

:::

---

## Clinical vs Patient-Reported Studies {.tabset}

::: {.panel-tabset}

### Clinical Studies

Clinical outcomes — measured by clinicians, laboratory tests, or objective performance assessments — tend to be less susceptible to expectation bias than self-report measures.
When a treatment shows an effect on clinical outcomes, it provides stronger evidence that something physiological is changing.

```{r}
#| label: forest-clinical
#| eval: !expr has_data
#| fig-alt: "Forest plot showing Hedges' g effect sizes for individual clinical studies, with a pooled diamond estimate at the bottom."
#| fig-cap: "Forest plot of treatment effects from clinical studies (Hedges' g)"
clinical <- effects[effects$reporting_type == "clinical", ]
if (nrow(clinical) >= 2) {
  es <- compute_effect_sizes(clinical, outcome_type = "continuous")
  fit <- run_meta(es)
  forest_gg(fit, title = "Treatment Effects — Clinical Studies")
}
```

```{r}
#| label: heterogeneity-clinical
#| eval: !expr has_data
if (nrow(clinical) >= 2) {
  het <- heterogeneity_summary(fit)
  cat(sprintf(
    "**Heterogeneity**: Q(%d) = %.1f, p = %.3f; I² = %.1f%%; τ² = %.3f\n",
    het$Q_df, het$Q, het$Q_p, het$I2, het$tau2
  ))
}
```

#### Subgroup: By PEM Status

Post-exertional malaise (PEM) is a cardinal feature of ME/CFS and common in Long COVID.
Trials that confirm PEM in their participants may be studying a biologically distinct population.

```{r}
#| label: subgroup-pem-clinical
#| eval: !expr has_data
#| fig-alt: "Forest plot of treatment effects stratified by whether the study population had confirmed post-exertional malaise."
#| fig-cap: "Treatment effects stratified by PEM status"
if (nrow(clinical) >= 2 && any(!is.na(clinical$pem_status))) {
  fit_sub <- run_subgroup_meta(es, moderator = "pem_status")
  cat(sprintf("Test of moderators: F = %.2f, p = %.3f\n", fit_sub$QM, fit_sub$QMp))
}
```

### Patient-Reported Studies

Patient-reported outcomes capture what matters most to the people living with these conditions: fatigue severity, functional capacity, quality of life.
They are also more vulnerable to placebo effects and unblinding, which is why we analyse them separately from clinical measures.

```{r}
#| label: forest-patient
#| eval: !expr has_data
#| fig-alt: "Forest plot showing Hedges' g effect sizes for individual patient-reported outcome studies, with a pooled diamond estimate."
#| fig-cap: "Forest plot of treatment effects from patient-reported studies"
patient <- effects[effects$reporting_type == "patient_reported", ]
if (nrow(patient) >= 2) {
  es_pat <- compute_effect_sizes(patient, outcome_type = "continuous")
  fit_pat <- run_meta(es_pat)
  forest_gg(fit_pat, title = "Treatment Effects — Patient-Reported Studies")
}
```

:::

## Study Quality {#sec-quality}

Risk of bias and GRADE ratings determine how much weight the pooled estimates deserve.
A meta-analysis built on studies with high risk of bias will produce a precise but potentially misleading number.

```{r}
#| label: rob-plot
#| eval: !expr nrow(quality) > 0
#| fig-alt: "Traffic light plot showing risk of bias judgements across multiple domains for each included treatment study."
#| fig-cap: "Risk of bias assessment for included treatment studies"
treatment_quality <- quality[quality$comparison != "", ]
if (nrow(treatment_quality) > 0) {
  rob_traffic_light(treatment_quality)
}
```

```{r}
#| label: grade-plot
#| eval: !expr nrow(quality) > 0
#| fig-alt: "Heatmap showing GRADE evidence certainty ratings for each treatment comparison, coloured from very low to high."
#| fig-cap: "GRADE evidence certainty ratings"
if (nrow(treatment_quality) > 0 && any(!is.na(treatment_quality$grade))) {
  grade_heatmap(treatment_quality)
}
```

## Publication Bias {#sec-pub-bias}

Funnel plot asymmetry can suggest that small studies with negative results are missing from the evidence base.
Asymmetry alone does not prove publication bias — it can also reflect genuine heterogeneity or differences in study quality — but it warrants caution.

```{r}
#| label: funnel-plot
#| eval: !expr has_data
#| fig-alt: "Funnel plot showing individual study effect sizes against their standard errors, with the pooled estimate marked as a vertical line."
#| fig-cap: "Funnel plot for treatment effectiveness studies"
if (exists("fit") && fit$k >= 5) {
  funnel_gg(fit)
}
```

## Sensitivity Analysis {#sec-sensitivity}

Leave-one-out analysis removes each study in turn and re-estimates the pooled effect.
If the overall result shifts substantially when a single study is dropped, that study is driving the conclusion and deserves closer scrutiny.

```{r}
#| label: leave-one-out
#| eval: !expr has_data
if (exists("fit") && fit$k >= 3) {
  l1o <- leave_one_out(fit)
  print(l1o)
}
```

::: {.callout-note collapse="true"}
## Data Status

```{r}
#| label: data-status
if (!has_data) {
  cat("No extracted treatment data yet. Run the search and extraction pipeline to populate this review.\n")
} else {
  cat(sprintf(
    "%d effect sizes from %d studies. Last updated: %s\n",
    nrow(effects),
    length(unique(effects$study_id)),
    max(studies$date_added[studies$search_topic == "treatment_effectiveness"], na.rm = TRUE)
  ))
}
```

This is a living review. New studies are incorporated as they are identified through automated search updates.
:::
