---
title: "Treatment Effectiveness"
---

```{r}
#| label: setup
#| include: false
source("../R/utils.R")
source("../R/meta-analysis.R")
source("../R/plots.R")

studies <- read_study_database()
effects <- read_extracted_data("treatment_effectiveness")
quality <- read_quality_ratings()
has_data <- nrow(effects) > 0
```

## Overview

This review examines the comparative effectiveness of interventions for Long COVID and ME/CFS, including:

- **Pacing / energy management** vs usual care or other active interventions
- **Graded exercise therapy (GET)** vs usual care or pacing
- **Cognitive behavioural therapy (CBT)** vs usual care
- **Pharmacological interventions** (LDN, antihistamines, beta-blockers, etc.)
- **Supplements** (CoQ10, NAD+, etc.)

Clinical and patient-reported outcome studies are analysed separately.

```{r}
#| label: included-studies
#| eval: !expr has_data
treatment_studies <- studies[studies$search_topic == "treatment_effectiveness" & studies$included, ]
cat(sprintf("**%d studies included** in treatment effectiveness review.\n", nrow(treatment_studies)))
```

## Clinical Studies {#sec-clinical}

```{r}
#| label: forest-clinical
#| eval: !expr has_data
#| fig-cap: "Forest plot of treatment effects from clinical studies (Hedges' g)"
clinical <- effects[effects$reporting_type == "clinical", ]
if (nrow(clinical) >= 2) {
  es <- compute_effect_sizes(clinical, outcome_type = "continuous")
  fit <- run_meta(es)
  forest_gg(fit, title = "Treatment Effects — Clinical Studies")
}
```

```{r}
#| label: heterogeneity-clinical
#| eval: !expr has_data
if (nrow(clinical) >= 2) {
  het <- heterogeneity_summary(fit)
  cat(sprintf(
    "**Heterogeneity**: Q(%d) = %.1f, p = %.3f; I² = %.1f%%; τ² = %.3f\n",
    het$Q_df, het$Q, het$Q_p, het$I2, het$tau2
  ))
}
```

### Subgroup: By PEM Status

```{r}
#| label: subgroup-pem-clinical
#| eval: !expr has_data
#| fig-cap: "Treatment effects stratified by PEM status"
if (nrow(clinical) >= 2 && any(!is.na(clinical$pem_status))) {
  fit_sub <- run_subgroup_meta(es, moderator = "pem_status")
  cat(sprintf("Test of moderators: F = %.2f, p = %.3f\n", fit_sub$QM, fit_sub$QMp))
}
```

## Patient-Reported Studies {#sec-patient-reported}

```{r}
#| label: forest-patient
#| eval: !expr has_data
#| fig-cap: "Forest plot of treatment effects from patient-reported studies"
patient <- effects[effects$reporting_type == "patient_reported", ]
if (nrow(patient) >= 2) {
  es_pat <- compute_effect_sizes(patient, outcome_type = "continuous")
  fit_pat <- run_meta(es_pat)
  forest_gg(fit_pat, title = "Treatment Effects — Patient-Reported Studies")
}
```

## Lay Summary {#sec-lay}

```{r}
#| label: lay-summary
#| eval: !expr has_data
#| fig-cap: "What patients and clinicians report about treatments"
cat("*This section provides accessible visualisations for non-specialist audiences.*\n\n")
```

## Study Quality {#sec-quality}

```{r}
#| label: rob-plot
#| eval: !expr nrow(quality) > 0
#| fig-cap: "Risk of bias assessment for included treatment studies"
treatment_quality <- quality[quality$comparison != "", ]
if (nrow(treatment_quality) > 0) {
  rob_traffic_light(treatment_quality)
}
```

```{r}
#| label: grade-plot
#| eval: !expr nrow(quality) > 0
#| fig-cap: "GRADE evidence certainty ratings"
if (nrow(treatment_quality) > 0 && any(!is.na(treatment_quality$grade))) {
  grade_heatmap(treatment_quality)
}
```

## Publication Bias {#sec-pub-bias}

```{r}
#| label: funnel-plot
#| eval: !expr has_data
#| fig-cap: "Funnel plot for treatment effectiveness studies"
if (exists("fit") && fit$k >= 5) {
  funnel_gg(fit)
}
```

## Sensitivity Analysis {#sec-sensitivity}

```{r}
#| label: leave-one-out
#| eval: !expr has_data
if (exists("fit") && fit$k >= 3) {
  l1o <- leave_one_out(fit)
  print(l1o)
}
```

::: {.callout-note}
## Data Status

```{r}
#| label: data-status
if (!has_data) {
  cat("No extracted treatment data yet. Run the search and extraction pipeline to populate this review.\n")
} else {
  cat(sprintf(
    "%d effect sizes from %d studies. Last updated: %s\n",
    nrow(effects),
    length(unique(effects$study_id)),
    max(studies$date_added[studies$search_topic == "treatment_effectiveness"], na.rm = TRUE)
  ))
}
```
:::
