---
title: "Treatment: CBT & Psychological Interventions"
aliases:
  - /reviews/cbt/
  - /reviews/psychological/
---

```{r}
#| label: setup
#| include: false
source(here::here("R/utils.R"))
source(here::here("R/meta-analysis.R"))
source(here::here("R/plots.R"))

studies <- read_study_database()
effects <- read_extracted_data("treatment_cbt_psych")
quality <- read_quality_ratings()
has_data <- nrow(effects) > 0
```

## Overview

This review covers cognitive behavioural therapy (CBT), graded exercise therapy (GET), acceptance and commitment therapy (ACT), the Lightning Process, mindfulness, and self-management interventions for Long COVID and ME/CFS.

```{r}
#| label: included-studies
#| results: asis
topic_studies <- studies |>
  dplyr::filter(
    !is.na(search_topic), search_topic == "treatment_cbt_psych",
    !is.na(llm_relevant), llm_relevant
  )
if (nrow(topic_studies) > 0) {
  n_rct <- sum(grepl("RCT|random", topic_studies$llm_study_type %||% "", ignore.case = TRUE), na.rm = TRUE)
  cli::cat_line(glue::glue("**{nrow(topic_studies)} relevant studies** ({n_rct} RCTs) identified."))
}
```

## Effect Sizes {#sec-effects}

```{r}
#| label: forest-data
#| include: false
forest_fit <- NULL
forest_h <- 6
if (has_data && all(c("mean_intervention", "sd_intervention", "n_intervention",
                      "mean_comparator", "sd_comparator", "n_comparator") %in% names(effects))) {
  complete <- effects |>
    dplyr::filter(
      !is.na(mean_intervention), !is.na(mean_comparator),
      !is.na(sd_intervention), !is.na(sd_comparator),
      !is.na(n_intervention), !is.na(n_comparator),
      n_intervention > 0, n_comparator > 0,
      sd_intervention > 0, sd_comparator > 0
    )
  if (nrow(complete) >= 2) {
    complete <- complete |>
      dplyr::rename(
        mean_treatment = mean_intervention,
        sd_treatment   = sd_intervention,
        n_treatment    = n_intervention,
        mean_control   = mean_comparator,
        sd_control     = sd_comparator,
        n_control      = n_comparator
      ) |>
      dplyr::mutate(study_label = study_id)
    es <- compute_effect_sizes(complete, outcome_type = "continuous")
    forest_fit <- run_meta(es)
    forest_h <- forest_height(forest_fit$k)
  }
}
```

```{r}
#| label: forest-plot
#| eval: !expr '!is.null(forest_fit)'
#| fig-height: !expr forest_h
#| fig-cap: "Effect sizes for psychological interventions (positive = benefit)"
forest_gg(forest_fit, title = "CBT & Psychological Interventions")
```

## Interventions {#sec-interventions}

```{r}
#| label: intervention-table
#| eval: !expr has_data
if (has_data && "intervention" %in% names(effects)) {
  int_counts <- effects |>
    dplyr::filter(!is.na(intervention)) |>
    dplyr::count(intervention) |>
    dplyr::arrange(dplyr::desc(n))
  if (nrow(int_counts) > 0) {
    names(int_counts) <- c("Intervention", "Effect Sizes")
    display_table(int_counts)
  }
}
```

## Outcome Measures {#sec-outcomes}

```{r}
#| label: outcome-table
#| eval: !expr has_data
if (has_data && "outcome_measure" %in% names(effects)) {
  out_counts <- effects |>
    dplyr::filter(!is.na(outcome_measure)) |>
    dplyr::count(outcome_measure) |>
    dplyr::arrange(dplyr::desc(n))
  if (nrow(out_counts) > 0) {
    names(out_counts) <- c("Outcome Measure", "N")
    display_table(out_counts)
  }
}
```

## PEM Assessment {#sec-pem}

```{r}
#| label: pem-status
#| results: asis
#| eval: !expr has_data
if (has_data && "pem_assessed" %in% names(effects)) {
  n_pem <- sum(effects$pem_assessed, na.rm = TRUE)
  n_total <- nrow(effects)
  pct <- round(n_pem / n_total * 100)
  cli::cat_line(glue::glue("PEM was assessed in **{n_pem}/{n_total}** ({pct}%) reported outcomes."))
}
```

## Risk of Bias {#sec-rob}

```{r}
#| label: rob-summary
#| eval: !expr nrow(quality) > 0
psych_quality <- quality |>
  dplyr::filter(!is.na(search_topic), search_topic == "treatment_cbt_psych",
                domain_name == "overall")
if (nrow(psych_quality) > 0) {
  rob_counts <- psych_quality |>
    dplyr::count(judgement)
  names(rob_counts) <- c("Overall RoB", "Studies")
  display_table(rob_counts)
}
```

## Adverse Events {#sec-safety}

```{r}
#| label: adverse-events
#| results: asis
#| eval: !expr has_data
if (has_data && "adverse_events_reported" %in% names(effects)) {
  n_ae <- sum(effects$adverse_events_reported, na.rm = TRUE)
  cli::cat_line(glue::glue("Adverse events reported in **{n_ae}/{nrow(effects)}** outcomes."))
}
```

::: {.callout-note}
## Data Status

```{r}
#| label: data-status
#| results: asis
if (!has_data) {
  cli::cat_line("No extracted data yet. Run the extraction pipeline to populate this review.")
} else {
  n_studies <- dplyr::n_distinct(effects$study_id)
  cli::cat_line(glue::glue("{nrow(effects)} effect sizes from {n_studies} studies."))
}
```
:::
