---
title: "Treatment: CBT & Psychological Interventions"
subtitle: "Evidence for behavioural and psychological approaches"
aliases:
  - /reviews/cbt/
  - /reviews/psychological/
---

```{r}
#| label: setup
#| include: false
library(targets)
tar_config_set(store = here::here("_targets"))
source(here::here("R/utils.R"))
source(here::here("R/meta-analysis.R"))
source(here::here("R/plots.R"))

studies <- tar_read(studies)
effects <- tar_read(cbt_effects)
quality <- tar_read(quality)
has_data <- nrow(effects) > 0
```

::: {.lead}
Psychological and behavioural interventions are among the most studied — and most debated — approaches for Long COVID and ME/CFS.
This review covers CBT, graded exercise therapy (GET), acceptance and commitment therapy (ACT), the Lightning Process, mindfulness, pacing programmes, and self-management interventions.
:::

```{r}
#| label: included-studies
#| results: asis
#| echo: false
counts <- tar_read(cbt_study_counts)
cat(sprintf('<div class="study-count"><span class="count-number">%d</span><span class="count-label">studies with<br>results</span></div>', counts$n_results))
cat(sprintf('<div class="study-count"><span class="count-number">%d</span><span class="count-label">randomised<br>controlled trials</span></div>', counts$n_rct))
cat(sprintf('<div class="study-count"><span class="count-number">%d</span><span class="count-label">registered<br>protocols</span></div>\n', counts$n_protocols))
```

The evidence base is large, but the picture is complicated.
Many trials use different outcome measures, different diagnostic criteria, and different definitions of what counts as "improvement."
The data below reflects that heterogeneity honestly.

```{r}
#| label: forest-data
#| include: false
forest_df <- tar_read(cbt_forest)
has_forest <- nrow(forest_df) >= 2
intervention_forest_h <- 6
domain_forest_h <- 6
if (has_forest && "intervention_category" %in% names(forest_df)) {
  n_int <- length(unique(forest_df$intervention_category[!is.na(forest_df$intervention_category)]))
  intervention_forest_h <- bar_height(n_int)
}
if (has_forest && "outcome_domain" %in% names(forest_df)) {
  n_dom <- length(unique(forest_df$outcome_domain[!is.na(forest_df$outcome_domain)]))
  domain_forest_h <- bar_height(n_dom)
}
```

::: {.at-a-glance}

## At a Glance {#sec-lay}

```{r}
#| label: lay-psych-lollipop
#| eval: !expr 'has_forest && "intervention_category" %in% names(forest_df)'
#| fig-height: 8
#| fig-alt: "Lollipop chart showing how much each psychological or behavioural treatment helps, with plain-language labels."
#| fig-cap: "How much does each treatment help? Each dot represents the pooled result from all studies of that treatment. Labels describe the size of the effect in plain language."
lay_lollipop(forest_df, group_col = "intervention_category",
             title = "Psychological & behavioural treatments: how much do they help?")
```

```{r}
#| label: lay-psych-verdict
#| eval: !expr 'has_forest && "intervention_category" %in% names(forest_df)'
#| fig-height: 8
#| fig-alt: "Traffic-light dot plot showing whether each psychological treatment is likely helpful, uncertain, or may not help."
#| fig-cap: "Does the treatment help? Green = evidence points towards benefit. Grey = we cannot tell from the available studies."
lay_treatment_summary(forest_df, group_col = "intervention_category",
                       title = "Psychological & behavioural treatments: does the evidence say they help?")
```

```{r}
#| label: lay-pem-waffle-psych
#| eval: !expr has_data
#| fig-height: 5
#| fig-alt: "Waffle chart showing that very few psychological treatment studies measured post-exertional malaise."
#| fig-cap: "Did these studies check for crashes after activity? Each square = 1% of study results. This matters because exercise-based treatments can trigger post-exertional malaise in susceptible patients, and if a study does not measure it, it cannot detect this harm."
if (has_data && "pem_assessed" %in% names(effects)) {
  n_pem <- sum(effects$pem_assessed, na.rm = TRUE)
  lay_waffle(n_pem, nrow(effects),
             yes_label = "Measured PEM",
             no_label = "Did not measure PEM",
             title = "How many studies checked for post-exertional crashes?")
}
```

:::

<hr class="section-divider">

## Effects by Intervention {#sec-effects-intervention}

Pooling every psychological intervention into a single number hides more than it reveals.
The plot below breaks the evidence apart by intervention category — each diamond is the pooled effect for one type of intervention, sized by the number of study–outcome pairs contributing.

```{r}
#| label: intervention-forest
#| eval: !expr 'has_forest && "intervention_category" %in% names(forest_df)'
#| fig-height: !expr intervention_forest_h
#| fig-alt: "Forest plot showing pooled effect sizes for each intervention category, with diamond size proportional to the number of contributing studies"
#| fig-cap: "Pooled effect sizes by intervention type. Each row is a separate random-effects meta-analysis. Larger diamonds indicate more contributing study–outcome pairs. Negative values favour the intervention over the comparator."
subgroup_forest(
  forest_df, group_col = "intervention_category",
  title = "Pooled effects by intervention type"
)
```

## Effects by Outcome Domain {#sec-effects-domain}

The same data, sliced differently.
Here the pooling is by what was measured rather than what was done.

```{r}
#| label: domain-forest
#| eval: !expr 'has_forest && "outcome_domain" %in% names(forest_df)'
#| fig-height: !expr domain_forest_h
#| fig-alt: "Forest plot showing pooled effect sizes for each outcome domain, with diamond size proportional to the number of contributing studies"
#| fig-cap: "Pooled effect sizes by outcome domain. Each row pools all interventions that measured that domain. Negative values indicate improvement."
subgroup_forest(
  forest_df, group_col = "outcome_domain",
  title = "Pooled effects by outcome domain"
)
```

## The PACE Trial Controversy {#sec-pace}

::: {.callout-warning}
## Why PACE matters for this review

The PACE trial (White et al., 2011; chalder2015 in our database) is the largest RCT of CBT and GET for ME/CFS (n = 641).
It contributes the largest individual effect sizes in this review and substantially influences the pooled estimates for both CBT and GET.

**The trial has been subject to extensive methodological criticism:**

- **Mid-trial outcome threshold changes**: the protocol-defined recovery thresholds for the primary outcome (Chalder Fatigue Scale) and key secondary outcome (SF-36 physical function) were [relaxed during the trial](https://doi.org/10.1186/s12916-019-1296-y), allowing participants to be classified as "recovered" while scoring within the entry criteria range.
- **No objective outcome measures**: primary outcomes were entirely self-reported; the only objective measure (actigraphy/6-minute walk test) showed no significant difference between groups.
- **Lack of blinding**: participants and therapists knew which treatment was received, creating expectation bias for subjective outcomes.
- **Reanalysis with original thresholds**: when independent researchers [reanalysed PACE data](https://doi.org/10.1186/s12916-018-1115-z) using the originally published recovery criteria, recovery rates dropped from 22% to 7% for CBT and from 22% to 4% for GET — not significantly different from the control group.
- **Regulatory response**: [NICE withdrew its recommendation of GET](https://www.nice.org.uk/guidance/ng206) for ME/CFS in 2021, citing the low quality of the evidence base that PACE had anchored.
:::

## Sensitivity: Excluding the PACE Trial {#sec-sensitivity}

The analysis below removes all PACE data to show how the pooled estimates change.

```{r}
#| label: sensitivity-no-pace
#| eval: !expr has_forest
#| results: asis
no_pace <- forest_df[forest_df$study_id != "chalder2015", ]
if (nrow(no_pace) >= 2) {
  pace_rows <- forest_df[forest_df$study_id == "chalder2015", ]

  cat(sprintf(
    "Removing %d effect sizes from the PACE trial leaves %d entries from %d studies.\n\n",
    nrow(pace_rows), nrow(no_pace), length(unique(no_pace$study_id))
  ))

  for (grp in c("CBT", "Graded Exercise Therapy")) {
    sub_full <- forest_df[!is.na(forest_df$intervention_category) &
                          forest_df$intervention_category == grp, ]
    sub_no_pace <- no_pace[!is.na(no_pace$intervention_category) &
                           no_pace$intervention_category == grp, ]
    if (nrow(sub_full) >= 2) {
      fit_full <- tryCatch(run_meta(sub_full), error = function(e) NULL)
      fit_np <- if (nrow(sub_no_pace) >= 2) tryCatch(run_meta(sub_no_pace), error = function(e) NULL) else NULL

      if (!is.null(fit_full)) {
        cat(sprintf("**%s** — with PACE: SMD = %.2f [%.2f, %.2f], k = %d",
            grp, coef(fit_full), fit_full$ci.lb, fit_full$ci.ub, fit_full$k))
        if (!is.null(fit_np)) {
          cat(sprintf(" | without PACE: SMD = %.2f [%.2f, %.2f], k = %d\n\n",
              coef(fit_np), fit_np$ci.lb, fit_np$ci.ub, fit_np$k))
        } else {
          cat(" | without PACE: insufficient data\n\n")
        }
      }
    }
  }
}
```

## Symptom Burden Evidence {#sec-symptom-burden}

::: {.key-finding}
Several studies report raw score changes on symptom burden measures (DSQ, Chalder Fatigue Scale, WSAS) rather than standardised effect sizes.
These cannot be pooled with the SMDs above but are important for assessing whether interventions worsen symptoms — particularly post-exertional malaise.
Positive values on symptom scales indicate **worsening**.
:::

```{r}
#| label: symptom-burden-data
#| include: false
symptom_df <- NULL
if (has_data && "effect_size_type" %in% names(effects)) {
  symptom_df <- effects |>
    dplyr::filter(
      effect_size_type == "mean_diff",
      !is.na(effect_size),
      grepl("CBT|GET|graded exercise|exercise|rehabilitation",
            intervention, ignore.case = TRUE)
    ) |>
    dplyr::mutate(
      study_label = dplyr::if_else(
        !is.na(outcome_measure),
        paste0(study_id, ": ", outcome_measure),
        study_id
      ),
      direction = dplyr::case_when(
        grepl("SF-36|SF-12|6.Minute|walk|EQ-5D|FACIT|strength|sit.to.stand",
              outcome_measure, ignore.case = TRUE) ~ "higher_better",
        TRUE ~ "lower_better"
      ),
      valence = dplyr::case_when(
        direction == "higher_better" & effect_size > 0 ~ "improved",
        direction == "higher_better" & effect_size < 0 ~ "worsened",
        direction == "lower_better" & effect_size < 0 ~ "improved",
        direction == "lower_better" & effect_size > 0 ~ "worsened",
        TRUE ~ "no change"
      )
    )
}
has_symptom <- !is.null(symptom_df) && nrow(symptom_df) > 0
```

```{r}
#| label: symptom-burden-table
#| eval: !expr has_symptom
symptom_summary <- symptom_df |>
  dplyr::select(study_label, intervention, outcome_measure, effect_size, valence) |>
  dplyr::arrange(valence, study_label)
names(symptom_summary) <- c("Study: Outcome", "Intervention", "Measure",
                             "Raw difference", "Direction")
display_table(symptom_summary)
```

## Interventions & Outcomes {.tabset}

::: {.panel-tabset}

### By Intervention

The table below groups individual intervention names into broader categories.
CBT and exercise-based approaches dominate the literature, but the category labels hide real diversity.

```{r}
#| label: intervention-table
#| eval: !expr has_data
if (has_data && "intervention_category" %in% names(effects)) {
  int_counts <- effects |>
    dplyr::filter(!is.na(intervention_category)) |>
    dplyr::count(intervention_category, sort = TRUE)
  if (nrow(int_counts) > 0) {
    names(int_counts) <- c("Intervention", "Effect Sizes")
    display_table(int_counts)
  }
}
```

### By Outcome Measure

Outcome measurement is one of the biggest sources of heterogeneity.
The same construct — fatigue, say — gets measured with the Chalder Fatigue Scale in one trial, the MFI-20 in another, and a simple VAS in a third.

```{r}
#| label: outcome-table
#| eval: !expr has_data
if (has_data && "outcome_measure" %in% names(effects)) {
  out_counts <- effects |>
    dplyr::filter(!is.na(outcome_measure)) |>
    dplyr::mutate(
      domain = dplyr::coalesce(outcome_domain, "other")
    ) |>
    dplyr::count(domain, outcome_measure, sort = TRUE)
  if (nrow(out_counts) > 0) {
    names(out_counts) <- c("Domain", "Outcome Measure", "N")
    display_table(out_counts)
  }
}
```

### By Domain

Rather than comparing individual instruments, we can group outcomes into broad domains to see where the evidence concentrates.

```{r}
#| label: domain-bar-data
#| include: false
domain_df <- NULL
domain_h <- 4
if (has_data && "outcome_domain" %in% names(effects)) {
  domain_df <- effects |>
    dplyr::filter(!is.na(outcome_domain)) |>
    dplyr::count(outcome_domain, sort = TRUE) |>
    dplyr::mutate(outcome_domain = factor(outcome_domain, levels = rev(outcome_domain)))
  domain_h <- bar_height(nrow(domain_df))
}
```

```{r}
#| label: domain-bar
#| eval: !expr '!is.null(domain_df) && nrow(domain_df) > 0'
#| fig-height: !expr domain_h
#| fig-alt: "Horizontal bar chart showing the number of extracted effect sizes per outcome domain, with fatigue and physical function having the most"
#| fig-cap: "Distribution of effect sizes across outcome domains."
ggplot(domain_df, aes(x = n, y = outcome_domain)) +
  geom_col(fill = cols_review["clinical"], width = 0.7) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Effect sizes", y = NULL) +
  theme_review
```

```{r}
#| label: domain-forest-data
#| include: false
domain_fits <- list()
domain_forest_h <- 6
if (has_data && all(c("mean_intervention", "sd_intervention", "n_intervention",
                      "mean_comparator", "sd_comparator", "n_comparator",
                      "outcome_domain") %in% names(effects))) {
  for (dom in unique(effects$outcome_domain[!is.na(effects$outcome_domain)])) {
    dom_data <- effects |>
      dplyr::filter(
        outcome_domain == dom,
        !is.na(mean_intervention), !is.na(mean_comparator),
        !is.na(sd_intervention), !is.na(sd_comparator),
        !is.na(n_intervention), !is.na(n_comparator),
        n_intervention > 0, n_comparator > 0,
        sd_intervention > 0, sd_comparator > 0
      ) |>
      dplyr::rename(
        mean_treatment = mean_intervention, sd_treatment = sd_intervention,
        n_treatment = n_intervention, mean_control = mean_comparator,
        sd_control = sd_comparator, n_control = n_comparator
      ) |>
      dplyr::mutate(study_label = paste0(study_id, ": ", outcome_measure))
    if (nrow(dom_data) >= 2) {
      es <- compute_effect_sizes(dom_data, outcome_type = "continuous")
      domain_fits[[dom]] <- run_meta(es)
    }
  }
  if (length(domain_fits) > 0) {
    total_k <- sum(vapply(domain_fits, \(f) f$k, integer(1)))
    domain_forest_h <- forest_height(total_k + length(domain_fits))
  }
}
```

```{r}
#| label: domain-summary-table
#| eval: !expr 'length(domain_fits) > 0'
domain_summary <- purrr::imap_dfr(domain_fits, \(fit, dom) {
  data.frame(
    Domain = dom,
    k = fit$k,
    SMD = round(coef(fit), 2),
    CI = paste0("[", round(fit$ci.lb, 2), ", ", round(fit$ci.ub, 2), "]"),
    p = formatC(fit$pval, format = "g", digits = 3),
    I2 = paste0(round(fit$I2, 1), "%")
  )
}) |> dplyr::arrange(SMD)
display_table(domain_summary)
```

:::

## PEM Assessment {#sec-pem}

::: {.key-finding}
Post-exertional malaise (PEM) is the hallmark symptom of ME/CFS and a core feature of Long COVID.
Whether a study assessed PEM matters because interventions that increase activity — like GET — can trigger PEM in susceptible patients.
A trial that does not measure PEM cannot detect this harm.
:::

```{r}
#| label: pem-status
#| results: asis
#| eval: !expr has_data
if (has_data && "pem_assessed" %in% names(effects)) {
  n_pem <- sum(effects$pem_assessed, na.rm = TRUE)
  n_total <- nrow(effects)
  pct <- round(n_pem / n_total * 100)
  cli::cat_line(glue::glue("PEM was assessed in **{n_pem}/{n_total}** ({pct}%) reported outcomes."))
}
```

## Risk of Bias {#sec-rob}

Risk of bias ratings follow the Cochrane RoB2 tool for RCTs and ROBINS-I for observational studies.

```{r}
#| label: rob-summary
#| eval: !expr nrow(quality) > 0
psych_quality <- quality |>
  dplyr::filter(!is.na(search_topic), search_topic == "treatment_cbt_psych",
                domain_name == "overall")
if (nrow(psych_quality) > 0) {
  rob_counts <- psych_quality |>
    dplyr::count(judgement)
  names(rob_counts) <- c("Overall RoB", "Studies")
  display_table(rob_counts)
}
```

```{r}
#| label: rob-forest-studies
#| eval: !expr 'has_forest && nrow(quality) > 0'
#| results: asis
forest_studies <- unique(forest_df$study_id)
forest_rob <- psych_quality[psych_quality$study_id %in% forest_studies, ]
if (nrow(forest_rob) > 0) {
  cat("\n**Risk of bias for studies in the forest plot:**\n\n")
  rob_forest_tbl <- forest_rob |>
    dplyr::select(study_id, judgement, rob_tool) |>
    dplyr::arrange(judgement)
  names(rob_forest_tbl) <- c("Study", "Overall RoB", "Tool")
  display_table(rob_forest_tbl)
} else {
  cat("\nNo quality ratings available for the studies in the forest plot.\n")
}
```

## GRADE Evidence Certainty {#sec-grade}

::: {.key-finding}
GRADE (Grading of Recommendations, Assessment, Development and Evaluation) rates the certainty of evidence across the full body of studies, considering risk of bias, inconsistency, indirectness, imprecision, and publication bias.
A "very low" rating means the true effect may be substantially different from the estimated effect.
:::

```{r}
#| label: grade-cbt
#| eval: !expr nrow(tar_read(grade)) > 0
grade_data <- tar_read(grade)
cbt_grade <- grade_data[grepl("Cbt|Psych|Exercise", grade_data$outcome, ignore.case = TRUE) |
                        grepl("Cbt|Psych|Exercise", grade_data$comparison %||% "", ignore.case = TRUE), ]
if (nrow(cbt_grade) > 0) {
  grade_tbl <- cbt_grade |>
    dplyr::select(
      dplyr::any_of(c("outcome", "certainty", "n_studies", "rationale"))
    ) |>
    dplyr::arrange(factor(certainty,
      levels = c("high", "moderate", "low", "very low")))
  display_table(grade_tbl)
}
```

```{r}
#| label: grade-note
#| eval: !expr nrow(tar_read(grade)) > 0
#| results: asis
if (nrow(cbt_grade) > 0) {
  n_vlow <- sum(cbt_grade$certainty == "very low", na.rm = TRUE)
  n_total <- nrow(cbt_grade)
  if (n_vlow > 0) {
    cat(sprintf(
      "\n**%d of %d** outcome domains are rated **very low** certainty — the lowest possible GRADE rating. For these outcomes, the direction and magnitude of the true effect remain highly uncertain.\n",
      n_vlow, n_total
    ))
  }
}
```

## Adverse Events {#sec-safety}

Adverse event reporting is inconsistent across trials.
Many studies do not mention adverse events at all, which does not mean none occurred — it means we cannot tell.

```{r}
#| label: adverse-events
#| results: asis
#| eval: !expr has_data
if (has_data && "adverse_events_reported" %in% names(effects)) {
  n_ae <- sum(effects$adverse_events_reported, na.rm = TRUE)
  cli::cat_line(glue::glue("Adverse events reported in **{n_ae}/{nrow(effects)}** outcomes."))
}
```

::: {.callout-note collapse="true"}
## Data Status

```{r}
#| label: data-status
#| results: asis
if (!has_data) {
  cli::cat_line("No extracted data yet. Run the extraction pipeline to populate this review.")
} else {
  n_studies <- dplyr::n_distinct(effects$study_id)
  cli::cat_line(glue::glue("{nrow(effects)} effect sizes from {n_studies} studies."))
}
```

This is a living review. Data is extracted automatically and updated as new evidence becomes available.
:::
